### 2025-03-07 project-proposal.qmd
Draft project proposal

# Opportunities for improvement in the care of adult patients with traumatic brain injury

## Introduction

Trauma - physical injury and the body's associated response - is a major cause of death and disability worldwide. Every year, over four million people die from trauma, and it is the top cause of burden of disease in young people. Traumatic brain injury is a major cause of death and disability in trauma patients. Trauma quality improvement programmes aim to identify opportunities for improvement in the care of trauma patients, and to implement changes to improve patient outcomes. The management of trauma patients once they reach hospital is very time sensitive and error prone, but little is known about potential opportunities for improvement among patients with traumatic brain injury.

## Specific aim

This study aims to describe opportunities for improvement among adult trauma patients with traumatic brain injury, and to assess factors associated with with opportunities for improvement in these patients.

## Work plan

Week 1-2: Finalise work- and project plan. Week 3: Write analysis plan. Week 4-10: Analysis. Week 11-12: Write results. Week 13-14: Write discussion. Week 15-20: Finalise, submit and present thesis.

## Material and methods

This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 14000 patients treated between 2012 and 2023. The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023. The outcome is opportunities for improvement, such as delayed management or technical errors, as identified by a multidisciplinary trauma quality improvement review board and recorded in the trauma care quality database. This project will link the two databases and assess what types of opportunities for improvement occurred in patients arriving with traumatic brain injury and use adjusted logistic regression to assess the association between different factors and the odds of opportunities for improvement. A 5% significance level and 95% confidence levels will be used.

## Collected data/samples

The outcome will be opportunities for improvement, coded as a nominal variable.

## Collected data

All data have already been collected and only need to be analyzed

## Type of project

Clinical study

## Ethical permit and review

Ethical permit is required and exists.

## Ethical review number or submission date

2021-02541, 2021-03531 and 2023-02975-02.

## Other students in related projects

### 2025-09-05 manuscript.Rmd
Start project

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->....

You can cite document here like this [@exampleKey9999]. Open the file
bibliography.bib to learn more.

Methods 
=======

Study design 10
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2025-09-06 manuscript.qmd
Test

---
title: "Manuscript"
date: today 
format: html
---

## Introduction

TBI TBI

### 2025-09-06 manuscript.qmd
test
Test

@@
 
-TBI TBI
+"Trauma - physical injury and the body's associated response - is a major cause of death and disability worldwide.
+Every year, over four million people die from trauma, and it is the top cause of the burden of disease in young people.
+Traumatic brain injury is a major cause of death and disability in trauma patients.
+Trauma quality improvement programmes aim to identify opportunities for improvement in the care of trauma patients and to implement changes to improve patient outcomes.
+The management of trauma patients once they reach the hospital is very time sensitive and error-prone, but little is known about potential opportunities for improvement among patients with traumatic brain injury."

### 2025-09-06 functions/Only_ofi_function.R
Figuring out fuction to only include pat with a +ve ofi

@@ -1,0 +1,9 @@
+library(dplyr)
+
+
+filter_ofi <- function(x) 
+{
+  dplyr::filter(x, !is.na(ofi) & ofi != "No") 
+  }
+
+ofi.data <- filter_ofi(merged.data)

### 2025-09-08 manuscript.qmd
Writing start of introduction

--- 2025-09-06 manuscript.qmd
@@
 
-TBI TBI
+Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
+Trauma, often synonymously used with serious injury, is defined as the physical injury and the body's associated response (1).
+Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages (2).

### 2025-09-08 project-plan.Rmd
Writing start of introduction

---
title: "Manuscript title here"
subtitle: "Add a Subtitle if Needed"
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
the console and you should only see headings -->

Introduction
============

*This is part A) in the project plan [sv: projektredogörelse]. *

*The background/introduction should summarise the scope of the problem,
what is known about the problem, what is not known, what specific
knowledge gap the study is supposed to fill and why this is
important. It should end be stating the aim of the study. The 
background should be 4-5 paragraphs long and each paragraph should be
between 3-5 sentences. Strive to make the paragraphs about the same
length.*

*bibliography.bib includes an example reference. Add additional
references in that file as bibtex entries and then cite in text using:
"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
https://www.doi2bib.org/. If you use Zotero you can simply export 
entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*

Methods
=======

*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

Project Update and Timeline
===========================

*This is part B) in the project plan [sv: projektredogörelse]. *

*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*

![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)

Back-up Plan
============

*This is part C) in the project plan [sv: projektredogörelse]*

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2025-09-08 functions/Only_ofi_function.R

@@ -3,7 +3,8 @@ library(dplyr)
 
 filter_ofi <- function(x) 
 {
-  dplyr::filter(x, !is.na(ofi) & ofi != "No") 
+  dplyr::filter(x, !is.na(ofi) & ofi != "No") #Filtering patients so that only those where ofi is +ve are left
   }
 
-ofi.data <- filter_ofi(merged.data)
+ofi.data <- filter_ofi(merged.data) #testing if function works 
+

### 2025-09-18 manuscript.qmd
writing intro
Writing the introduction and aim for the paper, needs to be shortened

--- 2025-09-06 manuscript.qmd
@@
 format: html
+# Shorten text to 4-5 paragraphs. Text rn is fot the project description
 ---
@@
 
-TBI TBI
+Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
+Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response.
+(1) Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
+(2)   # youuu crazyyy
+
+### Pathophysiology
+
+Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and constitutes a significant public health burden.
+(3) TBI is the result of an external force causing neurological impairments that may be either transient or permanent.
+(4) Furthermore, TBI should not be regarded as a single event but rather as a dynamic process.
+The initial trauma is followed by secondary pathophysiological cascades, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes.
+(5) As such, patients affected often require admission to the intensive care unit (ICU).
+
+### Management of TBI in the ICU 
+The management of TBI in the ICU is particularly challenging.
+Given the heterogeneity and dynamic nature of the condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia, hypercapnia, arterial hypotension, hyponatraemia and pyrexia.
+Equal attention must also be directed toward secondary intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure.
+(5–7) Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from optimal practice, thereby posing risks to both patient safety and overall quality of care.
+
+The intensive care of patients with TBI has been addressed in both guidelines and reviews.
+Multiple recent reviews and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central reference for modern management (5–9).
+According to the BTF, treatment should be initiated when intracranial pressure (ICP) exceeds 22 mmHg.
+Cerebral perfusion pressure (CPP) should be maintained between 60–70 mmHg.
+Prophylactic hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
+At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
+For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended.
+In cases where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and osmotherapy with mannitol or hypertonic saline.
+All this is always in conjunction with urgent neurosurgical evaluation.
+(Källa Karolinska handboken)
+
+### Trauma Quality Improvement
+
+Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors, deviations from established practice, and treatment delays, all of which may negatively influence outcomes.
+In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted benchmarking system designed to assess the quality of trauma care.
+Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
+Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16).
+Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error.
+Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16).
+Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+
+Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different aspects of trauma care.
+In 2023, Hussein A.
+et al.
+analysed patient- and process-related factors during trauma resuscitation and concluded that variables such as time to first CT and injury severity were strongly associated with OFI (17).
+Cao ES.
+et al.
+continued the focus on ICU cohorts, presenting that 9.9% ICU-admitted adult trauma patients experienced at least one OFI (1).
+Most recently, Szolnoky K et al.
+assessed OFI patterns in trauma cohorts and demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have preventable shortcomings in care (18).
+
+## Aim
+
+As such, this study aims to identify and describe opportunities for improvement in the care of adult trauma patients with TBI admitted to the ICU, and to examine patient- and process-related factors associated with their occurrence.

### 2025-09-19 manuscript.qmd
Writing the introduction, needs to be shortened and refined
intro
Intro writing
Adding comments to intro

@@
 ---
-
 ## Introduction
@@
 (1) Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
-(2)   # youuu crazyyy
+(2)
 
@@
 ### Management of TBI in the ICU
+
 The management of TBI in the ICU is particularly challenging.
@@
 All this is always in conjunction with urgent neurosurgical evaluation.
-(Källa Karolinska handboken)
+(Källa Karolinska handboken)  <!-- Make these two into one paragraph -->
 

### 2025-09-19 project-plan.qmd
Adding references using bibtex

--- 2025-09-19 manuscript.qmd
@@
 ---
-title: "Manuscript"
+title: "project-plan"
 date: today 
+bibliography: bibliography.bib
+csl: vancouver.csl
 format: html
-# Shorten text to 4-5 paragraphs. Text rn is fot the project description
 ---
+
+<!-- The project plan should be 3-4 pages long and written in
+Rmarkdown (like this document). Remove all comments in italic when you 
+use this document as a template for your own project plan. -->
+
+<!-- This is an R Markdown document. Markdown is a simple formatting
+syntax for authoring HTML, PDF, and MS Word documents. For more
+details on using R Markdown see http://rmarkdown.rstudio.com -->
+
+<!-- This is a comment and will not be present in the compiled
+document. Try "knitting" this document by clicking the knit button up
+to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
+the console and you should only see headings -->
+
 ## Introduction
 
+*This is part A) in the project plan [sv: projektredogörelse]. *
+
+*The background/introduction should summarise the scope of the problem,
+what is known about the problem, what is not known, what specific
+knowledge gap the study is supposed to fill and why this is
+important. It should end be stating the aim of the study. The 
+background should be 4-5 paragraphs long and each paragraph should be
+between 3-5 sentences. Strive to make the paragraphs about the same
+length.*
+
+*bibliography.bib includes an example reference. Add additional
+references in that file as bibtex entries and then cite in text using:
+"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
+https://www.doi2bib.org/. If you use Zotero you can simply export 
+entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
+Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*
+
+-------------------------------------------------
+
 Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
-Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response.
-(1) Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
+Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response "[@Szolnoky2025]"  (1).
+Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
 (2)
@@
 
-## Aim
+Methods
+=======
 
-As such, this study aims to identify and describe opportunities for improvement in the care of adult trauma patients with TBI admitted to the ICU, and to examine patient- and process-related factors associated with their occurrence.
+*Refer to the appropriate reporting guideline for details. If you are
+developing, updating or validating a clinical prediction model then
+use
+[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
+you are conducting an observational study, for example a cohort or
+case control study in which you assess associations between some
+exposure and an outcome then use
+[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*
+
+Project Update and Timeline
+===========================
+
+*This is part B) in the project plan [sv: projektredogörelse]. *
+
+*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*
+
+![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)
+
+Back-up Plan
+============
+
+*This is part C) in the project plan [sv: projektredogörelse]*
+
+References
+==========
+
+<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2025-09-24 manuscript.qmd
Writing method section in project plan, need to edit parts of introduction

@@
 format: html
-# Shorten text to 4-5 paragraphs. Text rn is fot the project description
+bibliography: bibliography.bib
+csl: vancouver.csl
+# Shorten text to 4-5 paragraphs. Text rn is for the project description
 ---
@@
 All this is always in conjunction with urgent neurosurgical evaluation.
-(Källa Karolinska handboken)  <!-- Make these two into one paragraph -->
+(Källa Karolinska handboken)  <!-- Make these two into one paragraph, remove local sources such as Karolinska handboken and keep it more generally applicable -->
 

### 2025-09-24 project-plan.qmd
Writing method section in project plan, need to edit parts of introduction

@@
 
-## Introduction
+Introduction
+============
+Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
+Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response [@Szolnoky2025].
+Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages [@Maerz2009].
 
-*This is part A) in the project plan [sv: projektredogörelse]. *
-
-*The background/introduction should summarise the scope of the problem,
-what is known about the problem, what is not known, what specific
-knowledge gap the study is supposed to fill and why this is
-important. It should end be stating the aim of the study. The 
-background should be 4-5 paragraphs long and each paragraph should be
-between 3-5 sentences. Strive to make the paragraphs about the same
-length.*
-
-*bibliography.bib includes an example reference. Add additional
-references in that file as bibtex entries and then cite in text using:
-"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
-https://www.doi2bib.org/. If you use Zotero you can simply export 
-entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
-Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*
-
--------------------------------------------------
-
-Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
-Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response "[@Szolnoky2025]"  (1).
-Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
-(2)
-
-### Pathophysiology
+Pathophysiology 
+------------
 
@@
 
-### Management of TBI in the ICU
+Management of TBI in the ICU 
+------------
 
@@
 
-### Trauma Quality Improvement
+Trauma Quality Improvement
+------------
 
@@
 =======
+
+Study Design
+-----------
+A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna (NKS).
+The trauma registry for NKS is a comprehensive trauma registry that reports to the Swedish National Trauma Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein Template).
+As such, the registry prospectively records patient demographics, prehospital and emergency department physiology, times of key interventions such as imaging, surgery and transfer, injury severity scores, and in-hospital outcome.
+
+The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events such as falls, traffic incidents, or other external violence.
+Since 2024 , patients have been included if a trauma team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS) greater than fifteen without trauma team activation.
+Patients transferred to a hospital within seven days of the traumatic event with a NISS greater than fifteen are also included.
+Exclusion criteria are cases where the only traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with protected identity, and patients declared dead on arrival.
+Since these inclusion and exclusion criteria have changed over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period (Källa SweTrau årsbok 2024).
+
+To identify opportunities for improvement, the trauma registry was linked to the hospital’s local trauma care quality database.
+Linkage was performed using the Swedish personal identity number, after which trauma registry variables and quality review outcomes were extracted for analysis of patient and process factors associated with opportunities for improvement.
+
+Setting
+---------
+Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced trauma care for both adults and children.
+According to the definition of the American College of Surgeons, Karolinska University Hospital in Solna serves as the designated Level 1 trauma centre for the greater metropolitan area of Stockholm.
+This designation requires around-the-clock availability of key services, including neurosurgery, general surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and consultants from other relevant specialities.
+(Källa   från American College of surgeons, plus källa  som säger NKS har alla dessa saker).
+Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units per the Svenska Föreningen för Anestesi och Intensivvård.
+(källa )
+
+Furthermore, the trauma registry includes about 14000 patients admitted between 2012 and 2023.
+The trauma care quality database is a subset of this registry, encompassing approximately 8,000 patients selected for review between 2014 and 2023.
+Screening for potential opportunities for improvement was implemented in 2013 for selected patient cases.
+Initially, case selection was based on a limited number of audit filters and the judgment of a small group of clinicians directly involved in trauma care.
+Since 2017, a standardised set of audit filters has been consistently applied, and all trauma patients admitted to the hospital are screened for potential opportunities for improvement.
+The process has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and comprehensive review.
+Patients flagged through the audit filters, and therefore have potential OFI, undergo discussion in structured multidisciplinary morbidity and mortality conferences, where representatives from relevant specialties and professions involved in trauma care are involved.
+The discussion focuses on establishing whether there were any preventable events or deviations from standards of practice.
+The final classification of whether an opportunity for improvement is present is determined by consensus among the peer-review group.
+As such, subjectivity is minimised and ensures that decisions reflect a collective judgement.
+Whether or not an OFI is identified, the decision of the review is recorded in the trauma care quality database, ensuring systematic documentation of positive findings and cases without deviations.
+
+Participants
+------------
+Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been screened for opportunity for improvements.
+Furthermore, TBI was defined as (a GCS score of ≤8 combined with an Abbreviated Injury Scale ≥3 for the head region ….
+) Patients were eligible irrespective of level of care and were not restricted to intensive care admissions.
+
+Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality review processes that differ from those for adults .
+Also excluded were patients declared dead on arrival, since assessment of opportunities for improvement in these cases is not relevant to the aim of this study.
+
+Variables
+---------
+Outcome
+
+The primary outcome of this study was the presence of one or more opportunities for improvement during the management of patients with TBI.
+As mentioned, OFIs were adjudicated after a screening process with audit filters and a subsequent review in multidisciplinary morbidity-and-mortality conferences.
+Following the peer review, the result of each case discussion was recorded in the trauma care quality register.
+Moreover, the outcome was presented as a binary variable, “Yes - At least one OFI identified” and “No - No OFI identified”.
+
+Patient factors and processes
+
+The variables chosen to be analysed for this study were those from the trauma registry, based on the locally used audit filters, standard epidemiological factors and the factors registered in the Swedish National Trauma Registry.
+These could be divided into categorical and continuous variables.
+
+Certain continuous variables were retained in their continuous form for analysis, but others were categorised into clinically meaningful groups to facilitate descriptive statistics and subgroup analyses.
+The continuous variables include…
+And those who were categorised were....
+
+Furthermore, the categorical variables include…
+
+Statistical methods
+----------------
 

### 2025-09-25 project-plan.qmd
Adding general management about TBI in intro

@@
 
+Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological stabilisation and diagnostics in order for secondary injuries to be limited.
+Protection of the airway in patients with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased mortality (6–9).
+Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
+Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for severely injured patients (10,11).
+Haemostatic strategies in the early stages, such as administering tranexamic acid within three hours of injury, can reduce head injury-related mortality (12).
+After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic and intracranial insults becomes the primary focus.
+
 The management of TBI in the ICU is particularly challenging.
@@
 Prophylactic hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
-At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
+
+(At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
 For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended.
@@
 All this is always in conjunction with urgent neurosurgical evaluation.
-(Källa Karolinska handboken)  <!-- Make these two into one paragraph -->
+(Källa Karolinska handboken))  <!-- Remove?
+-->
 
@@
 assessed OFI patterns in trauma cohorts and demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have preventable shortcomings in care (18).
+<!-- Rewrite into a more general rationale paragraph-->
 

### 2025-09-25 project-plan.qmd
Addin comments into participants
Adding comments

@@
 ============
-Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
-Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response [@Szolnoky2025].
-Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages [@Maerz2009].
+Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest
+members of society. Trauma, often synonymously used with serious injury, is defined as a physical injury and the
+body's associated response [@Szolnoky2025]. Furthermore, trauma remains a leading cause of death and disability
+worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading
+cause of death across all ages [@Maerz2009].
 
@@
 
-Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and constitutes a significant public health burden.
-(3) TBI is the result of an external force causing neurological impairments that may be either transient or permanent.
-(4) Furthermore, TBI should not be regarded as a single event but rather as a dynamic process.
-The initial trauma is followed by secondary pathophysiological cascades, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes.
-(5) As such, patients affected often require admission to the intensive care unit (ICU).
+Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
+constitutes a significant public health burden. (3) TBI is the result of an external force causing neurological
+impairments that may be either transient or permanent.  (4) Furthermore, TBI should not be regarded as a single event
+but rather as a dynamic process. The initial trauma is followed by secondary pathophysiological cascades, including
+ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes. (5)
+As such, patients affected often require admission to the intensive care unit (ICU).
 
-Management of TBI in the ICU 
+Management of TBI 
 ------------
 
-Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological stabilisation and diagnostics in order for secondary injuries to be limited.
-Protection of the airway in patients with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased mortality (6–9).
-Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
-Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for severely injured patients (10,11).
-Haemostatic strategies in the early stages, such as administering tranexamic acid within three hours of injury, can reduce head injury-related mortality (12).
-After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic and intracranial insults becomes the primary focus.
+Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological
+stabilisation and diagnostics in order for secondary injuries to be limited. Protection of the airway in patients
+with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all
+prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased
+mortality (6–9).  Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
+Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality
+advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for
+severely injured patients (10,11). Haemostatic strategies in the early stages, such as administering tranexamic acid
+within three hours of injury, can reduce mortality (12).
+After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic
+and intracranial insults becomes the primary focus.
 
-The management of TBI in the ICU is particularly challenging.
-Given the heterogeneity and dynamic nature of the condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia, hypercapnia, arterial hypotension, hyponatraemia and pyrexia.
-Equal attention must also be directed toward secondary intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure.
-(5–7) Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from optimal practice, thereby posing risks to both patient safety and overall quality of care.
+The management of TBI in the ICU is particularly challenging. Given the heterogeneity and dynamic nature of the
+condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia,
+hypercapnia, arterial hypotension, hyponatraemia and pyrexia. Equal attention must also be directed toward secondary
+intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure. (5–7)
+Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from
+optimal practice, thereby posing risks to both patient safety and overall quality of care.
 
-The intensive care of patients with TBI has been addressed in both guidelines and reviews.
-Multiple recent reviews and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central reference for modern management (5–9).
+The intensive care of patients with TBI has been addressed in both guidelines and reviews. Multiple recent reviews
+and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central
+reference for modern management (5–9).
 According to the BTF, treatment should be initiated when intracranial pressure (ICP) exceeds 22 mmHg.
 Cerebral perfusion pressure (CPP) should be maintained between 60–70 mmHg.
-Prophylactic hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
+Prophylactic
+hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
 
 (At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
-For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended.
-In cases where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and osmotherapy with mannitol or hypertonic saline.
-All this is always in conjunction with urgent neurosurgical evaluation.
-(Källa Karolinska handboken))  <!-- Remove?
--->
+For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT
+imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended. In cases
+where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and
+osmotherapy with mannitol or hypertonic saline. All this is always in conjunction with urgent neurosurgical
+evaluation. (Källa Karolinska handboken))  <!-- Remove paragraph? -->
 
@@
 
-Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors, deviations from established practice, and treatment delays, all of which may negatively influence outcomes.
-In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted benchmarking system designed to assess the quality of trauma care.
-Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
-Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16).
-Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error.
-Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16).
-Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors,
+deviations from established practice, and treatment delays, all of which may negatively influence outcomes. In
+response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted
+benchmarking system designed to assess the quality of trauma care. Through audit filters and peer review, it
+identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11). Previous
+studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in
+imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid
+management (12–16). Importantly, some of these findings were observed in high-performing level I trauma centres,
+underscoring that even mature systems remain vulnerable to error. Furthermore, injuries to the central nervous system
+account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such
+as the Arkansas Trauma Registry (16). Yet, while previous research has described OFI in trauma resuscitation and in
+general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI
+specifically among patients with TBI. <!-- Add a part after ACS TQIP using WHO as a reference to TQIP-->
 
-Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different aspects of trauma care.
-In 2023, Hussein A.
-et al.
-analysed patient- and process-related factors during trauma resuscitation and concluded that variables such as time to first CT and injury severity were strongly associated with OFI (17).
-Cao ES.
-et al.
-continued the focus on ICU cohorts, presenting that 9.9% ICU-admitted adult trauma patients experienced at least one OFI (1).
-Most recently, Szolnoky K et al.
-assessed OFI patterns in trauma cohorts and demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have preventable shortcomings in care (18).
-<!-- Rewrite into a more general rationale paragraph-->
+Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different
+aspects of trauma care. In 2023, Hussein A. et al. analysed patient- and process-related factors during trauma
+resuscitation and concluded that variables such as time to first CT and injury severity were strongly associated with
+OFI (17). Cao ES. et al. continued the focus on ICU cohorts, presenting that 9.9% ICU-admitted adult trauma patients
+experienced at least one OFI (1). Most recently, Szolnoky K et al. assessed OFI patterns in trauma cohorts and
+demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have
+preventable shortcomings in care (18). <!-- Rewrite into a more general rationale paragraph-->
 
@@
 -----------
-A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna (NKS).
-The trauma registry for NKS is a comprehensive trauma registry that reports to the Swedish National Trauma Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein Template).
-As such, the registry prospectively records patient demographics, prehospital and emergency department physiology, times of key interventions such as imaging, surgery and transfer, injury severity scores, and in-hospital outcome.
+A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna
+(NKS). The trauma registry for NKS is a comprehensive trauma registry that reports to the Swedish National Trauma
+Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein
+Template). As such, the registry prospectively records patient demographics, prehospital and emergency department
+physiology, times of key interventions such as imaging, surgery and transfer, injury severity scores, and in-hospital outcome.
 
-The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events such as falls, traffic incidents, or other external violence.
-Since 2024 , patients have been included if a trauma team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS) greater than fifteen without trauma team activation.
-Patients transferred to a hospital within seven days of the traumatic event with a NISS greater than fifteen are also included.
-Exclusion criteria are cases where the only traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with protected identity, and patients declared dead on arrival.
-Since these inclusion and exclusion criteria have changed over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period (Källa SweTrau årsbok 2024).
+The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events
+such as falls, traffic incidents, or other external violence. Since 2024 , patients have been included if a trauma
+team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS)
+greater than fifteen without trauma team activation. Patients transferred to a hospital within seven days of the
+traumatic event with a NISS greater than fifteen are also included. Exclusion criteria are cases where the only
+traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with
+protected identity, and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed
+over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period
+(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data is up to 2023 -->
 
-To identify opportunities for improvement, the trauma registry was linked to the hospital’s local trauma care quality database.
-Linkage was performed using the Swedish personal identity number, after which trauma registry variables and quality review outcomes were extracted for analysis of patient and process factors associated with opportunities for improvement.
+To identify opportunities for improvement, the trauma registry was linked to the hospital’s local trauma care quality
+database. Linkage was performed using the Swedish personal identity number, after which trauma registry variables and
+quality review outcomes were extracted for analysis of patient and process factors associated with opportunities for
+improvement.
 
@@
 ---------
-Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced trauma care for both adults and children.
-According to the definition of the American College of Surgeons, Karolinska University Hospital in Solna serves as the designated Level 1 trauma centre for the greater metropolitan area of Stockholm.
-This designation requires around-the-clock availability of key services, including neurosurgery, general surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and consultants from other relevant specialities.
-(Källa   från American College of surgeons, plus källa  som säger NKS har alla dessa saker).
-Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units per the Svenska Föreningen för Anestesi och Intensivvård.
-(källa )
+Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced
+trauma care for both adults and children. According to the definition of the American College of Surgeons, Karolinska
+University Hospital in Solna serves as the designated Level 1 trauma centre for the greater metropolitan area of
+Stockholm. This designation requires around-the-clock availability of key services, including neurosurgery, general
+surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and
+consultants from other relevant specialities. (Källa   från American College of surgeons, plus källa  som säger NKS
+har alla dessa saker). Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units
+per the Svenska Föreningen för Anestesi och Intensivvård. (källa )
 
-Furthermore, the trauma registry includes about 14000 patients admitted between 2012 and 2023.
-The trauma care quality database is a subset of this registry, encompassing approximately 8,000 patients selected for review between 2014 and 2023.
-Screening for potential opportunities for improvement was implemented in 2013 for selected patient cases.
-Initially, case selection was based on a limited number of audit filters and the judgment of a small group of clinicians directly involved in trauma care.
-Since 2017, a standardised set of audit filters has been consistently applied, and all trauma patients admitted to the hospital are screened for potential opportunities for improvement.
-The process has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and comprehensive review.
-Patients flagged through the audit filters, and therefore have potential OFI, undergo discussion in structured multidisciplinary morbidity and mortality conferences, where representatives from relevant specialties and professions involved in trauma care are involved.
-The discussion focuses on establishing whether there were any preventable events or deviations from standards of practice.
-The final classification of whether an opportunity for improvement is present is determined by consensus among the peer-review group.
-As such, subjectivity is minimised and ensures that decisions reflect a collective judgement.
-Whether or not an OFI is identified, the decision of the review is recorded in the trauma care quality database, ensuring systematic documentation of positive findings and cases without deviations.
+Furthermore, the trauma registry includes about 14000 patients admitted between 2012 and 2023. The trauma care
+quality database is a subset of this registry, encompassing approximately 8,000 patients selected for review between
+2014 and 2023. Screening for potential opportunities for improvement was implemented in 2013 for selected patient
+cases. Initially, case selection was based on a limited number of audit filters and the judgment of a small group of
+clinicians directly involved in trauma care. Since 2017, a standardised set of audit filters has been consistently
+applied, and all trauma patients admitted to the hospital are screened for potential opportunities for improvement.
+The process has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and
+comprehensive review. Patients flagged through the audit filters, and therefore have potential OFI, undergo
+discussion in structured multidisciplinary morbidity and mortality conferences, where representatives from relevant
+specialties and professions involved in trauma care are involved.  The discussion focuses on establishing whether
+there were any preventable events or deviations from standards of practice. The final classification of whether an
+opportunity for improvement is present is determined by consensus among the peer-review group. As such, subjectivity
+is minimised and ensures that decisions reflect a collective judgement. Whether or not an OFI is identified, the
+decision of the review is recorded in the trauma care quality database, ensuring systematic documentation of positive
+findings and cases without deviations.
 
@@
 ------------
-Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been screened for opportunity for improvements.
-Furthermore, TBI was defined as (a GCS score of ≤8 combined with an Abbreviated Injury Scale ≥3 for the head region ….
-) Patients were eligible irrespective of level of care and were not restricted to intensive care admissions.
+Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
+2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as (a GCS score of ≤8
+combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level <!-- Definition found in Elins paper, same applies?-->
+of care and were not restricted to intensive care admissions.
 
-Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality review processes that differ from those for adults .
-Also excluded were patients declared dead on arrival, since assessment of opportunities for improvement in these cases is not relevant to the aim of this study.
+Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not fins source-->
+review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since
+assessment of opportunities for improvement in these cases is not relevant to the aim of this study.
 
@@
 
-The primary outcome of this study was the presence of one or more opportunities for improvement during the management of patients with TBI.
-As mentioned, OFIs were adjudicated after a screening process with audit filters and a subsequent review in multidisciplinary morbidity-and-mortality conferences.
-Following the peer review, the result of each case discussion was recorded in the trauma care quality register.
-Moreover, the outcome was presented as a binary variable, “Yes - At least one OFI identified” and “No - No OFI identified”.
+The primary outcome of this study was the presence of one or more opportunities for improvement during the management
+of patients with TBI. As mentioned, OFIs were determined after a screening process with audit filters and a
+subsequent review in multidisciplinary morbidity-and-mortality conferences. Following the peer review, the result of
+each case discussion was recorded in the trauma care quality register and was presented as a binary variable, “Yes -
+At least one OFI identified” and “No - No OFI identified”.
 
@@
 
-The variables chosen to be analysed for this study were those from the trauma registry, based on the locally used audit filters, standard epidemiological factors and the factors registered in the Swedish National Trauma Registry.
+The variables chosen to be analysed for this study were those from the trauma registry, based on the locally used
+audit filters, standard epidemiological factors and the factors registered in the Swedish National Trauma Registry.
 These could be divided into categorical and continuous variables.
 
-Certain continuous variables were retained in their continuous form for analysis, but others were categorised into clinically meaningful groups to facilitate descriptive statistics and subgroup analyses.
-The continuous variables include…
+Certain continuous variables were retained in their continuous form for analysis, but others were categorised into
+clinically meaningful groups to facilitate descriptive statistics and subgroup analyses.  The continuous variables
+include…  
 And those who were categorised were....
@@
 Furthermore, the categorical variables include…
+<!-- Choose variables, not sure which too chose-->
 
@@
 ----------------
+<!-- unsure -->
 
-*Refer to the appropriate reporting guideline for details. If you are
+<!-- *Refer to the appropriate reporting guideline for details. If you are
 developing, updating or validating a clinical prediction model then
@@
 exposure and an outcome then use
-[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*
+[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).* -->
 

### 2025-09-25 project-plan.qmd
Adding the Aim

@@
 over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period
-(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data is up to 2023 -->
+(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data used is up to 2023 -->
 
@@
 har alla dessa saker). Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units
-per the Svenska Föreningen för Anestesi och Intensivvård. (källa )
+per the Svenska Föreningen för Anestesi och Intensivvård. (källa)
 
@@
 
-Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not fins source-->
+Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not find source-->
 review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since
@@
 ----------------
-<!-- unsure -->
+
+For continuous variables, medians and interquartile ranges (25th-75th  percentiles) are used. Categorical variables
+are presented using the number of patients and their corresponding percentages of the cohort. <!-- If multivariable regression and anlysis will be used, how will we handle missing data? -->
+
+Logistic regression was used to examine the association between patient and process-related factors and OFI. Results
+are reported as Odds Ratio (OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold
+for statistical significance. To ensure objectivity, all statistical analyses were initially performed on synthetic
+data before being applied to the combined trauma registry and trauma care quality database. Statistical analyses were
+conducted in R version x.
+
+<!-- unsure how the statistical model will work, previous papers mention multivariable logistic regression and one Uni/bivariable regression. How does this work? -->
 

### 2025-09-25 project-plan.qmd
Adding section about WHO:s TQIP guidelines into intro

@@
 the console and you should only see headings -->
+
+<!--NEED TO ADD SOURCES IN TEXT-->
 
@@
 benchmarking system designed to assess the quality of trauma care. Through audit filters and peer review, it
-identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11). Previous
-studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in
-imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid
-management (12–16). Importantly, some of these findings were observed in high-performing level I trauma centres,
-underscoring that even mature systems remain vulnerable to error. Furthermore, injuries to the central nervous system
-account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such
-as the Arkansas Trauma Registry (16). Yet, while previous research has described OFI in trauma resuscitation and in
-general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI
-specifically among patients with TBI. <!-- Add a part after ACS TQIP using WHO as a reference to TQIP-->
+identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
+The WHO, however, insist that the trauma quality improvement programmes must be a continuous, multidisciplinary process.
+This
+entails audit filters, structured morbidity and mortality and preventable death case reviews in order to find
+opportunities for improvement (OFI). They also discuss that the formulation of corrective strategies or action plans
+in response to identified OFI:s is essential to trauma quality improvement programmes. Furthermore, the WHO
+emphasises that TQIP must ensure that corrective measures addressing identified OFI result in tangible and verifiable
+improvements in outcomes, thereby confirming that deficiencies are not only recognised but also effectively resolved. (Källa WHO) <!-- add WHO source -->
+
+Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16).
+Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error.
+Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16).
+Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+<!-- Add a part after ACS TQIP using WHO as a reference to TQIP-->
 
@@
 preventable shortcomings in care (18). <!-- Rewrite into a more general rationale paragraph-->
+
+Aim
+=====
+As such, this study aims to identify and describe opportunities for improvement in the care of adult trauma patients
+with TBI, and to examine patient- and process-related factors associated with their occurrence.
 
@@
 Furthermore, the categorical variables include…
-<!-- Choose variables, not sure which too chose-->
+<!-- Choose variables, not sure which ones to chose-->
 
@@
 For continuous variables, medians and interquartile ranges (25th-75th  percentiles) are used. Categorical variables
-are presented using the number of patients and their corresponding percentages of the cohort. <!-- If multivariable regression and anlysis will be used, how will we handle missing data? -->
+are presented using the number of patients and their corresponding percentages of the cohort.
+<!-- If multivariable regression analysis will be used, how will we handle missing data from other variables in the same patient?
+-->
 
-Logistic regression was used to examine the association between patient and process-related factors and OFI. Results
-are reported as Odds Ratio (OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold
-for statistical significance. To ensure objectivity, all statistical analyses were initially performed on synthetic
-data before being applied to the combined trauma registry and trauma care quality database. Statistical analyses were
-conducted in R version x.
-
-<!-- unsure how the statistical model will work, previous papers mention multivariable logistic regression and one Uni/bivariable regression. How does this work? -->
+Logistic regression was used to examine the association between patient and process-related factors and OFI.
+Bivariable logistic regression to assess unadjusted associations between patient and process factors and OFI was
+initially performed. Variables of clinical relevance and those significant in bivariable analyses were then entered
+into a multivariable logistic regression model to estimate adjusted associations. Results are reported as Odds Ratio
+(OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold for statistical
+significance. To ensure objectivity, all statistical analyses were initially performed on synthetic data before being
+applied to the combined trauma registry and trauma care quality database. Statistical analyses were conducted in R
+version x. <!-- I assume we will do both bivariable and multivariable regression? -->
 

### 2025-09-26 project-plan.qmd

@@
 Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
-<!-- Add a part after ACS TQIP using WHO as a reference to TQIP-->
 
-Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different
+TBI remains a challenge within modern trauma systems, not only due to its contribution to mortality but also due to
+the magnifying impact on the outcome because of small process deviations. Delays in imaging, inconsistent airway
+management, and lapses in haemodynamic measures are all examples that can result in irreversible damage. Although
+TQIP have been developed to identify and correct such lapses, with the WHO stressing the importance of linking OFIs
+to corrective actions and verification of bettered outcomes, most research has been done on mixed trauma cohorts.
+This, on the other hand, leaves TBI-specific vulnerabilities undercharacterised. Recent work by Szolnoky et al.
+suggests that, despite overall declines in OFI, patients with TBI remain at risk of preventable shortcomings,
+underscoring the need for TBI-specific quality improvement efforts (23).
+
+<!--Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different
 aspects of trauma care. In 2023, Hussein A. et al. analysed patient- and process-related factors during trauma
@@
 demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have
-preventable shortcomings in care (18). <!-- Rewrite into a more general rationale paragraph-->
+preventable shortcomings in care (18).-->
 
@@
 are presented using the number of patients and their corresponding percentages of the cohort.
-<!-- If multivariable regression analysis will be used, how will we handle missing data from other variables in the same patient?
+<!-- If multivariable regression analysis will be used, how will we handle missing data from other variables for the same patient?
 -->

### 2025-09-26 project-plan.qmd
Comment on project plan

@@
 Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
-constitutes a significant public health burden. (3) TBI is the result of an external force causing neurological
-impairments that may be either transient or permanent.  (4) Furthermore, TBI should not be regarded as a single event
-but rather as a dynamic process. The initial trauma is followed by secondary pathophysiological cascades, including
+constitutes a significant public health burden.(3) TBI is the result of an external force causing neurological
+impairments that may be either transient or permanent.(4) Furthermore, TBI should not be regarded as a single event
+but rather as a dynamic process. The initial trauma **often triggers a secondary pathophysiological cascade**, including
 ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes. (5)
-As such, patients affected often require admission to the intensive care unit (ICU).
+As such, patients **with TBI** often require admission to the intensive care unit (ICU).
 
@@
 ------------
+
+<!-- MGW: Consider discussing the management of trauma first and then the management of TBI, as many of the things you write below also apply to non-TBI trauma patients -->
 
@@
 
+<!-- MGW: Try to find a balance when it comes to the level of detail that you use. The specific ICD and CPP targets feel a bit too specific. -->
+
 (At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
@@
 osmotherapy with mannitol or hypertonic saline. All this is always in conjunction with urgent neurosurgical
-evaluation. (Källa Karolinska handboken))  <!-- Remove paragraph? -->
+evaluation. (Källa Karolinska handboken))  <!-- Remove paragraph? MGW: Yes, I think so.-->
 
@@
 response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted
-benchmarking system designed to assess the quality of trauma care. Through audit filters and peer review, it
+benchmarking system designed to assess the quality of trauma care <!-- MGW: I would mention ACS TQIP as one example of a TQIP and refer to the WHO guidelines as more general.
+I suggest that you start by introducing the concept of TQIP by referring to the WHO guidelines and then mention ACS TQIP as one example -->.
+Through audit filters and peer review, it
 identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
@@
 Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+
+<!-- MGW: I think this paragraph is good, but I suggest adding a subheading like "Rationale for this study" or something like that. -->
 
@@
 =====
-As such, this study aims to identify and describe opportunities for improvement in the care of adult trauma patients
+This study aims to identify and describe opportunities for improvement in the care of adult trauma patients
 with TBI, and to examine patient- and process-related factors associated with their occurrence.
@@
 A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna
-(NKS). The trauma registry for NKS is a comprehensive trauma registry that reports to the Swedish National Trauma
+(NKS). The trauma registry for NKS is a comprehensive trauma registry that **is part of** the Swedish National Trauma
 Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein
@@
 
-The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events
-such as falls, traffic incidents, or other external violence. Since 2024 , patients have been included if a trauma
+SweTrau includes patients of all ages who have sustained serious injuries from traumatic events
+such as falls, traffic incidents, or other external violence. Since 2024<!-- MGW: Has this changed over the years? Otherwise I suggest that you remove the year. -->, patients are included if a trauma
 team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS)
@@
 Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced
-trauma care for both adults and children. According to the definition of the American College of Surgeons, Karolinska
-University Hospital in Solna serves as the designated Level 1 trauma centre for the greater metropolitan area of
-Stockholm. This designation requires around-the-clock availability of key services, including neurosurgery, general
+trauma care for both adults and children<!-- MGW: Do you have a source for this? -->. The trauma center has around-the-clock availability of key services, including neurosurgery, general
 surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and
@@
 har alla dessa saker). Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units
-per the Svenska Föreningen för Anestesi och Intensivvård. (källa)
+according to the **Swedish Society of Anaesthesiology and Intensive Care Medicine**. (källa)
+
+<!-- The numbers below belong to the results section, so they should be moved there. -->
 
@@
 ------------
-Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
-2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as (a GCS score of ≤8
-combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level <!-- Definition found in Elins paper, same applies?-->
-of care and were not restricted to intensive care admissions.
+All adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
+2023 who also had been screened for opportunity for improvements **were included in this study**. Furthermore, TBI was defined as (a GCS score of ≤8
+combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level <!-- Definition found in Elins paper, same applies?--><!-- Yes -->
+of care.
 
@@
 ---------
-Outcome
+
+### Outcome
 
@@
 
-Patient factors and processes
+### Patient factors and processes
 
@@
 <!-- Choose variables, not sure which ones to chose--> 
+<!-- MGW: Start with the ones in Husseins and Elins papers-->
 
@@
 
+<!-- MGW: Regardless of the analysis method, we need to handle missing data somehow.
+You can use complete case analysis, which means that you only include cases with complete data for all variables of interest.
+-->
+
 Logistic regression was used to examine the association between patient and process-related factors and OFI.
@@
 initially performed. Variables of clinical relevance and those significant in bivariable analyses were then entered
-into a multivariable logistic regression model to estimate adjusted associations. Results are reported as Odds Ratio
+into a multivariable logistic regression model to estimate adjusted associations.
+<!-- MGW: I think you should decide which ones to include in the multivariable model from the start, regardless of their significance in the bivariable analysis --> Results are reported as Odds Ratio
 (OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold for statistical

### 2025-09-26 project-plan.qmd
Revision based on feedback

@@
 Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
-constitutes a significant public health burden.(3) TBI is the result of an external force causing neurological
-impairments that may be either transient or permanent.(4) Furthermore, TBI should not be regarded as a single event
-but rather as a dynamic process. The initial trauma **often triggers a secondary pathophysiological cascade**, including
-ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes. (5)
-As such, patients **with TBI** often require admission to the intensive care unit (ICU).
+constitutes a significant public health burden (3). TBI is the result of an external force causing neurological
+impairments that may be either transient or permanent (4). Furthermore, TBI should not be regarded as a single event
+but rather as a dynamic process.
+The initial trauma often triggers a secondary pathophysiological cascade, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes (5).
+As such, patients with TBI often require admission to the intensive care unit (ICU).
 
@@
 
-(At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
-For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT
-imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended. In cases
-where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and
-osmotherapy with mannitol or hypertonic saline. All this is always in conjunction with urgent neurosurgical
-evaluation. (Källa Karolinska handboken))  <!-- Remove paragraph? MGW: Yes, I think so.-->
-
 Trauma Quality Improvement
@@
 
-<!--Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different
-aspects of trauma care. In 2023, Hussein A. et al. analysed patient- and process-related factors during trauma
-resuscitation and concluded that variables such as time to first CT and injury severity were strongly associated with
-OFI (17). Cao ES. et al. continued the focus on ICU cohorts, presenting that 9.9% ICU-admitted adult trauma patients
-experienced at least one OFI (1). Most recently, Szolnoky K et al. assessed OFI patterns in trauma cohorts and
-demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have
-preventable shortcomings in care (18).-->
-
 Aim
@@
 A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna
-(NKS). The trauma registry for NKS is a comprehensive trauma registry that **is part of** the Swedish National Trauma
+(NKS). The trauma registry for NKS is a comprehensive trauma registry that is part of the Swedish National Trauma
 Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein
@@
 ------------
-All adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
-2023 who also had been screened for opportunity for improvements **were included in this study**. Furthermore, TBI was defined as (a GCS score of ≤8
-combined with an Abbreviated Injury Scale ≥3 for the head region ….  ) Patients were eligible irrespective of level <!-- Definition found in Elins paper, same applies?--><!-- Yes -->
-of care.
+All adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been
+screened for opportunity for improvements were included in this study. Furthermore, TBI was defined as a GCS score of
+≤8 combined with an Abbreviated Injury Scale ≥3 for the head region. Patients were eligible irrespective of level of
+care.
 
@@
 applied to the combined trauma registry and trauma care quality database. Statistical analyses were conducted in R
-version x. <!-- I assume we will do both bivariable and multivariable regression? -->
+version x.
 

### 2025-10-02 project-plan.qmd
Adding sources + inserting feedback from coordinator

@@
 ---
-
-<!-- The project plan should be 3-4 pages long and written in
-Rmarkdown (like this document). Remove all comments in italic when you 
-use this document as a template for your own project plan. -->
-
-<!-- This is an R Markdown document. Markdown is a simple formatting
-syntax for authoring HTML, PDF, and MS Word documents. For more
-details on using R Markdown see http://rmarkdown.rstudio.com -->
-
-<!-- This is a comment and will not be present in the compiled
-document. Try "knitting" this document by clicking the knit button up
-to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
-the console and you should only see headings -->
-
-<!--NEED TO ADD SOURCES IN TEXT-->
-
 Introduction
@@
 members of society. Trauma, often synonymously used with serious injury, is defined as a physical injury and the
-body's associated response [@Szolnoky2025]. Furthermore, trauma remains a leading cause of death and disability
+body's associated response [@Cao]. Furthermore, trauma remains a leading cause of death and disability
 worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading
@@
 Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
-constitutes a significant public health burden (3). TBI is the result of an external force causing neurological
-impairments that may be either transient or permanent (4). Furthermore, TBI should not be regarded as a single event
-but rather as a dynamic process.
-The initial trauma often triggers a secondary pathophysiological cascade, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes (5).
-As such, patients with TBI often require admission to the intensive care unit (ICU).
+constitutes a significant public health burden [@Guan2023]. TBI is the result of an external force causing
+neurological impairments that may be either transient or permanent [@Capizzi2020]. Furthermore, TBI should not be
+regarded as a single event but rather as a dynamic process. The initial trauma often triggers a secondary
+pathophysiological cascade, including ischemia, oedema, and metabolic disturbances, that develop over time and
+critically influence long-term outcomes [@Robba2025]. As such, patients with TBI often require admission to the
+intensive care unit (ICU).
 
@@
 prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased
-mortality (6–9). Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
-Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality
-advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for
-severely injured patients (10,11). Haemostatic strategies in the early stages, such as administering tranexamic acid
-within three hours of injury, can reduce mortality (12).
-After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic
-and intracranial insults becomes the primary focus.
+mortality  [@Carney2017; @Stocchetti2017; @Chesnut1993; @Spaite2017]. Non-contrast head CT should be performed
+without delay to identify surgically remediable lesions. Depending on mechanism and physiology, whole-body CT may be
+used selectively, as randomised data show no mortality advantage over targeted imaging, while observational work
+underscores the benefit of immediate CT pathways for severely injured patients [@Sierink2016; @Yamamoto2023].
+Haemostatic strategies in the early stages, such as administering tranexamic acid within three hours of injury, can
+reduce mortality [@Crash-32019]. After this initial phase, depending on the patient's condition, management
+transitions to other areas of the hospital, where prevention of secondary systemic and intracranial insults becomes
+the primary focus.
 
@@
 hypercapnia, arterial hypotension, hyponatraemia and pyrexia. Equal attention must also be directed toward secondary
-intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure. (5–7)
-Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from
-optimal practice, thereby posing risks to both patient safety and overall quality of care.
+intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure
+[@Robba2025; @Stocchetti2017; @Slot2025]. Although these complexities are often anticipated, they render ICU care
+especially vulnerable to deviations from optimal practice, thereby posing risks to both patient safety and overall
+quality of care.
 
@@
 and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central
-reference for modern management (5–9).
-According to the BTF, treatment should be initiated when intracranial pressure (ICP) exceeds 22 mmHg.
-Cerebral perfusion pressure (CPP) should be maintained between 60–70 mmHg.
-Prophylactic
-hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
-
-<!-- MGW: Try to find a balance when it comes to the level of detail that you use. The specific ICD and CPP targets feel a bit too specific. -->
+reference for modern management [@Robba2025; @Stocchetti2017; @Slot2025; @Wiles2022; @Dash2018]. The BTF advises
+treatment strategies to lower ICP and how to maintain cerebral perfusion pressure (CPP) within optimal ranges. It
+also discusses what practices to avoid, such as prophylactic hyperventilation within the first 24 hours and
+corticosteroid usage.
 
@@
 Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors,
-deviations from established practice, and treatment delays, all of which may negatively influence outcomes. In
-response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted
-benchmarking system designed to assess the quality of trauma care <!-- MGW: I would mention ACS TQIP as one example of a TQIP and refer to the WHO guidelines as more general.
-I suggest that you start by introducing the concept of TQIP by referring to the WHO guidelines and then mention ACS TQIP as one example -->.
-Through audit filters and peer review, it
-identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
+deviations from established practice, and treatment delays, all of which may negatively influence outcomes <!-- CP: Reference-->.
+In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP),
+a risk-adjusted benchmarking system designed to assess the quality of trauma care <!-- MGW: I would mention ACS TQIP
+as one example of a TQIP and refer to the WHO guidelines as more general. I suggest that you start by introducing the
+concept of TQIP by referring to the WHO guidelines and then mention ACS TQIP as one example -->.
+Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality [@Al-thani2023; @Shafi2009].
 The WHO, however, insist that the trauma quality improvement programmes must be a continuous, multidisciplinary process.
-This
-entails audit filters, structured morbidity and mortality and preventable death case reviews in order to find
-opportunities for improvement (OFI). They also discuss that the formulation of corrective strategies or action plans
-in response to identified OFI:s is essential to trauma quality improvement programmes. Furthermore, the WHO
-emphasises that TQIP must ensure that corrective measures addressing identified OFI result in tangible and verifiable
-improvements in outcomes, thereby confirming that deficiencies are not only recognised but also effectively resolved. (Källa WHO) <!-- add WHO source -->
+This entails audit filters, structured morbidity and mortality and preventable death case reviews in order to find opportunities for improvement (OFI).
+They also discuss that the formulation of corrective strategies or action plans in response to identified OFI:s is essential to trauma quality improvement programmes.
+Furthermore, the WHO emphasises that TQIP must ensure that corrective measures addressing identified OFI result in tangible and verifiable improvements in outcomes, thereby confirming that deficiencies are not only recognised but also effectively resolved [@world_health_organization_guidelines_2009].
 
-Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16).
-Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error.
-Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16).
-Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays
+in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and
+fluid management [@Sanddal2011; @Gruen2006; @Alao2024; @Teixeira2007; @Mabry2024]. Importantly, some of these
+findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain
+vulnerable to error. Furthermore, injuries to the central nervous system, like TBI, account for a substantial
+proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma
+Registry [@Mabry2024]. Yet, while previous research has described OFI in trauma resuscitation and in general ICU
+trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among
+patients with TBI.
 
-<!-- MGW: I think this paragraph is good, but I suggest adding a subheading like "Rationale for this study" or something like that. -->
+Rationale
+------------
 
@@
 to corrective actions and verification of bettered outcomes, most research has been done on mixed trauma cohorts.
-This, on the other hand, leaves TBI-specific vulnerabilities undercharacterised. Recent work by Szolnoky et al.
+This, on the other hand, leaves TBI-specific vulnerabilities undercharacterised. Recent work by *Szolnoky et al.*
 suggests that, despite overall declines in OFI, patients with TBI remain at risk of preventable shortcomings,
-underscoring the need for TBI-specific quality improvement efforts (23).
+underscoring the need for TBI-specific quality improvement efforts [@Szolnoky2025]. <!--CP: Therefore we hypothesis ze that is possible to identify new OFis related to TBI?? 
+CP: See if something like that can be a hypothesis -->
 
@@
 =====
-This study aims to identify and describe opportunities for improvement in the care of adult trauma patients
-with TBI, and to examine patient- and process-related factors associated with their occurrence.
+This study aims to identify and describe OFI in the care of adult trauma patients with TBI, and to examine patient- and process-related factors associated with their occurrence.
+<!-- CP: You need to phrase these aims in a more concrete way: this what this course demands..see also next comment E.g.
+(but for your supervisor to help with) Identify specific OFI by doing what?
+I mean , an identification require that certain parameters will be used and confirmed
+If too long M&M will describe these parameters)” -->
+
+<!-- CP: Pls describe this aim in a Yes/No question e..g determine whether the identified OFI are associated with patient and process related factors. Ask for your supervisor to help revising -->
+
+<!--CP: Can you explain more concretely What this means? Occurrence of OPI?-->
 
@@
 (NKS). The trauma registry for NKS is a comprehensive trauma registry that is part of the Swedish National Trauma
-Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein
-Template). As such, the registry prospectively records patient demographics, prehospital and emergency department
-physiology, times of key interventions such as imaging, surgery and transfer, injury severity scores, and in-hospital outcome.
+Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma [@Ringdal2008]. As
+such, the registry records patient demographics, prehospital and emergency department physiology, times of key
+interventions such as imaging, injury severity scores, and in-hospital outcomes.
 
-SweTrau includes patients of all ages who have sustained serious injuries from traumatic events
-such as falls, traffic incidents, or other external violence. Since 2024<!-- MGW: Has this changed over the years? Otherwise I suggest that you remove the year. -->, patients are included if a trauma
-team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS)
-greater than fifteen without trauma team activation. Patients transferred to a hospital within seven days of the
-traumatic event with a NISS greater than fifteen are also included. Exclusion criteria are cases where the only
-traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with
-protected identity, and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed
-over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period
-(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data used is up to 2023 -->
+The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events
+such as falls, traffic incidents, or other external violence. Patients have been included if a trauma team activation
+has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS) greater than fifteen
+without trauma team activation. Patients transferred to a hospital within seven days of the traumatic event with a
+NISS greater than fifteen are also included. Exclusion criteria are cases where the only traumatic condition is
+chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with protected identity,
+and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed over time, patients
+in this study will be filtered to ensure they fulfil the criteria relevant to the study period [@Holmberg_2023].
 
@@
 ---------
-Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced
-trauma care for both adults and children<!-- MGW: Do you have a source for this? -->. The trauma center has around-the-clock availability of key services, including neurosurgery, general
-surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and
-consultants from other relevant specialities. (Källa   från American College of surgeons, plus källa  som säger NKS
-har alla dessa saker). Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units
-according to the **Swedish Society of Anaesthesiology and Intensive Care Medicine**. (källa)
+Karolinska University Hospital in Solna houses Trauma Centre Karolinska, Sweden’s largest provider of advanced trauma
+care for both adults and children. According to the definition of the American College of Surgeons, NKS serves as the
+designated Level 1 trauma centre for the greater metropolitan area of Stockholm. This designation requires
+around-the-clock availability of key services, including neurosurgery, general surgery, orthopaedic surgery,
+anaesthesia and intensive care, as well as diagnostic and interventional radiology and consultants from other
+relevant specialities [@American_college_of_surgeons_resources_2022; @Granstrom2018]. Additionally, the ICU units in
+NKS are all recognised as Level 3 units according to the Swedish Society of Anaesthesiology and Intensive Care
+Medicine [@Swedish_society_of_anaesthesiology_and_intensive_care_medicine_riktlinjer_2024].
 
-<!-- The numbers below belong to the results section, so they should be moved there. -->
-
-Furthermore, the trauma registry includes about 14000 patients admitted between 2012 and 2023. The trauma care
-quality database is a subset of this registry, encompassing approximately 8,000 patients selected for review between
-2014 and 2023. Screening for potential opportunities for improvement was implemented in 2013 for selected patient
-cases. Initially, case selection was based on a limited number of audit filters and the judgment of a small group of
-clinicians directly involved in trauma care. Since 2017, a standardised set of audit filters has been consistently
-applied, and all trauma patients admitted to the hospital are screened for potential opportunities for improvement.
-The process has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and
+Furthermore, the trauma registry covers trauma patients admitted between 2012 and 2023. The trauma care quality
+database is a subset of this registry and contains patients who have been selected for review between 2014 and 2023.
+Screening for potential opportunities for improvement was implemented in 2013 for selected patient cases. Initially,
+case selection was based on a limited number of audit filters and the judgment of a small group of clinicians
+directly involved in trauma care. Since 2017, a standardised set of audit filters has been consistently applied, and
+all trauma patients admitted to the hospital are screened for potential opportunities for improvement. The process
+has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and
 comprehensive review. Patients flagged through the audit filters, and therefore have potential OFI, undergo
@@
 ------------
-All adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been
-screened for opportunity for improvements were included in this study. Furthermore, TBI was defined as a GCS score of
-≤8 combined with an Abbreviated Injury Scale ≥3 for the head region. Patients were eligible irrespective of level of
-care.
+Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
+2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as a Glasgow Coma
+Scale (GCS) score of ≤8 combined with an Abbreviated Injury Scale (AIS) ≥3 for the head region. Patients were
+eligible irrespective of level of care and were not restricted to intensive care admissions.
 
-Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not find source-->
-review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since
+Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality
+review processes that differ from those for adults. Also excluded were patients declared dead on arrival, since
 assessment of opportunities for improvement in these cases is not relevant to the aim of this study.
@@
 
-The primary outcome of this study was the presence of one or more opportunities for improvement during the management
-of patients with TBI. As mentioned, OFIs were determined after a screening process with audit filters and a
-subsequent review in multidisciplinary morbidity-and-mortality conferences. Following the peer review, the result of
+The primary outcome of this study was the presence of one or more OFIs during the management <!-- CP: You probably have all needed info in this text.
+However, Pls describe this section in relation to Aim 1 and 2.
+The text here needs to be stringently matched with the two aims-->
+of patients with TBI. As mentioned, OFIs were adjudicated after a screening process with audit filters and a
+subsequent review in multidisciplinary morbidity and mortality conferences. Following the peer review, the result of
 each case discussion was recorded in the trauma care quality register and was presented as a binary variable, “Yes -
@@
 
-Certain continuous variables were retained in their continuous form for analysis, but others were categorised into
-clinically meaningful groups to facilitate descriptive statistics and subgroup analyses.  The continuous variables
-include…  
-And those who were categorised were....
+The categorical variables include sex, on-call times defined as hours outside of 08:00 - 17:00, or Saturday and
+Sunday, hospital care level ordered from the emergency department to the highest being the ICU, intubation status,
+mechanism of injury, ASA class, duration of mechanical ventilation and ICU length stay. The latter two are
+categorised into different groups that represent a certain number of days.
 
-Furthermore, the categorical variables include…
-<!-- Choose variables, not sure which ones to chose--> 
-<!-- MGW: Start with the ones in Husseins and Elins papers-->
+The continuous variables include age, Injury Severity Score (ISS), Revised Trauma Score (RTS), systolic blood
+pressure (SBP), respiratory rate (RR) and time to first CT in minutes. The Revised Trauma Score was calculated using
+GCS, systolic blood pressure, and respiratory rate, each coded on a scale from 0 to 4 and combined with published
+coefficients [@Champion1989]. Furthermore, certain continuous variables were retained in their continuous form for
+analysis, but others were categorised into clinically meaningful groups to facilitate descriptive statistics and
+subgroup analyses.
 
-Statistical methods
+Statistical methods. <!-- CP: Pls recall aim 1 and aim 2 and develop the below text in relation to these 2 aims: see criteria-->
 ----------------
@@
 For continuous variables, medians and interquartile ranges (25th-75th  percentiles) are used. Categorical variables
-are presented using the number of patients and their corresponding percentages of the cohort.
-<!-- If multivariable regression analysis will be used, how will we handle missing data from other variables for the same patient?
--->
-
-<!-- MGW: Regardless of the analysis method, we need to handle missing data somehow.
-You can use complete case analysis, which means that you only include cases with complete data for all variables of interest.
--->
+are presented using the number of patients and their corresponding percentages of the cohort. Complete case analysis
+will be used; as such, only cases with complete data for all variables of interest will be used.
 
@@
 Bivariable logistic regression to assess unadjusted associations between patient and process factors and OFI was
-initially performed. Variables of clinical relevance and those significant in bivariable analyses were then entered
-into a multivariable logistic regression model to estimate adjusted associations.
-<!-- MGW: I think you should decide which ones to include in the multivariable model from the start, regardless of their significance in the bivariable analysis --> Results are reported as Odds Ratio
-(OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold for statistical
-significance. To ensure objectivity, all statistical analyses were initially performed on synthetic data before being
-applied to the combined trauma registry and trauma care quality database. Statistical analyses were conducted in R
-version x.
+initially performed. To estimate adjusted associations, these variables were entered into a multivariable logistic
+regression model. Results are reported as Odds Ratio (OR) with 95% confidence interval (CI), and a p-value of <0.05
+was used as the threshold for statistical significance. To ensure objectivity, all statistical analyses were
+initially performed on synthetic data before being applied to the combined trauma registry and trauma care quality
+database. Statistical analyses were conducted in R.
 
-<!-- *Refer to the appropriate reporting guideline for details. If you are
-developing, updating or validating a clinical prediction model then
-use
-[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
-you are conducting an observational study, for example a cohort or
-case control study in which you assess associations between some
-exposure and an outcome then use
-[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).* -->
+Ethical considerations
+=======
+
+This is a register-based study with no intervention and thus entailed no risk of burden nor physical harm or contact
+with patients. Rather, it involves secondary use of routinely collected clinical data. The principal ethical risk was
+the risk of privacy and integrity intrusion, especially during the linkage of the trauma registry and the local
+trauma care quality database before pseudonymization. Also, all results are reported on an aggregate level to
+minimise the possibility of reidentification. In line with regular practice for registry studies based on already
+collected data, informed consent was not obtained from participants. The risk of privacy intrusion was judged to
+increase if individual research subjects were to be contacted.
+
+The risk of integrity breach was weighed against the potential benefits the project can provide regarding health and
+was judged favourable. Although there is no direct benefit for the individuals whose data was used, the societal
+value is substantial because identifying opportunities for improvement in traumatic brain injury care can inform
+corrective actions, enhance patient safety, and reduce morbidity and mortality. The Swedish Ethical Review Authority
+granted ethical approval with the DNRs being: 2021-02541, 2021-03531 and 2023-02975-02.
 
@@
 
-*This is part B) in the project plan [sv: projektredogörelse]. *
+Data collection has already been done using two established registers at Karolinska University Hospital, Solna. These
+are the trauma registry (SweTrau) with approximately 14,000 patients between 2012–2023, and the trauma care quality
+database, which contains around 8,000 patients between 2014–2023. The latter includes the nominable variable of Yes
+or No OFI identified during multidisciplinary morbidity and mortality conferences.
 
-*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*
+Variables: Patient and process factors are analysed
 
-![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)
+-	Categorical: sex, arrival outside office hours, hospital care level, intubation status, mechanism of injury, ASA
+class, ventilation duration, and ICU length of stay.
 
+-	Continuous: age, Injury Severity Score, Revised Trauma Score, systolic blood pressure, respiratory rate, and time to CT. Some continuous variables will be categorised.
+
+Analysis:
+
+-	Results will be presented in tables and figures with for eg. Means, interquartile ranges, counts and percentages.
+
+-	Associations between factors and OFI will be examined using bivariable and multivariable logistic regression
+
+This will be presented as OR with 95% CI. Analyses will be complete-case and conducted in R.
+
+![**Gantt timeline for the project**](assets/project-gantt.png)
 Back-up Plan
@@
 
-*This is part C) in the project plan [sv: projektredogörelse]*
+If unforeseen issues arise during the project, and as data has already been collected, the project will be adjusted by doing the following:
+
+- **Reduce the dataset scope**  
+  Instead of analysing all patients with TBI in the register between 2012–2023, the reserve plan would focus on a shorter time period and therefore reduce the number of patients analysed.
+
+- **Reduce the variable set**  
+  If analysis of the full range of patient- and process-level factors proves too difficult to do within the timeframe, the analysis will prioritise some variables over others and reduce the number of variables.
+
+- **Alternative analysis**  
+  If multivariable regression is not possible or proves too difficult, the reserve plan is to perform simpler analysis, such as simple/bivariable analysis, therefore only presenting unadjusted associations.
 

### 2025-10-02 project-plan.qmd
Adding comments from coordinator

@@
 ---
+<!-- CP: The report is interesting, very promising and also well written. 
+Effort is needed to further improve the report by structuring it in relation to the aims.
+Aims also to be improved according to the course i.e.
+challenging a scientific question (possibly having a hypothesis) in a yes/no manner: see comments
+I might have missed some points in the report not being a clinician working on this matter but I think that you can improve the structure of the Aims and M&M a bit more. -->
+
 Introduction

### 2025-10-05 manuscript.qmd
Adding initial functions for the regression model

--- 2025-09-18 manuscript.qmd
@@
 ---
-title: "Manuscript"
-date: today 
-format: html
-# Shorten text to 4-5 paragraphs. Text rn is fot the project description
+title: "Manuscript title here"
+output: html_document
+author: "Your name here"
+bibliography: bibliography.bib
+csl: vancouver.csl
 ---
 
-## Introduction
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
-Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest members of society.
-Trauma, often synonymously used with serious injury, is defined as a physical injury and the body's associated response.
-(1) Furthermore, trauma remains a leading cause of death and disability worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading cause of death across all ages.
-(2)   # youuu crazyyy
+<!-- This is an R Markdown document. Markdown is a simple formatting
+syntax for authoring HTML, PDF, and MS Word documents. For more
+details on using R Markdown see http://rmarkdown.rstudio.com -->
 
-### Pathophysiology
+<!-- This is a comment and will not be present in the compiled
+document. Try "knitting" this document by clicking the knit button up
+to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
+the console and you should only see headings -->
 
-Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and constitutes a significant public health burden.
-(3) TBI is the result of an external force causing neurological impairments that may be either transient or permanent.
-(4) Furthermore, TBI should not be regarded as a single event but rather as a dynamic process.
-The initial trauma is followed by secondary pathophysiological cascades, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes.
-(5) As such, patients affected often require admission to the intensive care unit (ICU).
+<!-- Structure and most comments are from the Strengthening the
+reporting of observational studies in epidemiology (STROBE) statement,
+see https://www.strobe-statement.org/. For more explanations see the
+paper Strengthening the Reporting of Observational Studies in
+Epidemiology (STROBE):explanation and elaboration.  -->
 
-### Management of TBI in the ICU 
-The management of TBI in the ICU is particularly challenging.
-Given the heterogeneity and dynamic nature of the condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia, hypercapnia, arterial hypotension, hyponatraemia and pyrexia.
-Equal attention must also be directed toward secondary intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure.
-(5–7) Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from optimal practice, thereby posing risks to both patient safety and overall quality of care.
+<!-- Indicate the study’s design with a commonly used term in the
+title, for example: "Associations between resuscitation interventions
+and opportunities for improvement in adult trauma patients: A cohort
+study" -->
 
-The intensive care of patients with TBI has been addressed in both guidelines and reviews.
-Multiple recent reviews and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central reference for modern management (5–9).
-According to the BTF, treatment should be initiated when intracranial pressure (ICP) exceeds 22 mmHg.
-Cerebral perfusion pressure (CPP) should be maintained between 60–70 mmHg.
-Prophylactic hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.
-At Karolinska University Hospital, the setting for this study, similar principles are seen in the Traumahandboken.
-For severe TBI, defined as GCS < 8, early airway protection with intubation in unconscious patients, rapid CT imaging, timely administration of tranexamic acid, and early neurosurgical consultation are recommended.
-In cases where there are signs of impending herniation, the handbook advises measures such as controlled hyperventilation and osmotherapy with mannitol or hypertonic saline.
-All this is always in conjunction with urgent neurosurgical evaluation.
-(Källa Karolinska handboken)
+Abstract
+========
 
-### Trauma Quality Improvement
+<!-- Provide in the abstract an informative and balanced summary of
+what was done and what was found. Not more than 300 words. -->
 
-Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors, deviations from established practice, and treatment delays, all of which may negatively influence outcomes.
-In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted benchmarking system designed to assess the quality of trauma care.
-Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11).
-Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16).
-Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error.
-Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16).
-Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.
+Background
+----------
 
-Previous papers at Karolinska University Hospital have contributed to this field by describing OFIs across different aspects of trauma care.
-In 2023, Hussein A.
-et al.
-analysed patient- and process-related factors during trauma resuscitation and concluded that variables such as time to first CT and injury severity were strongly associated with OFI (17).
-Cao ES.
-et al.
-continued the focus on ICU cohorts, presenting that 9.9% ICU-admitted adult trauma patients experienced at least one OFI (1).
-Most recently, Szolnoky K et al.
-assessed OFI patterns in trauma cohorts and demonstrated that, although overall incidence declined over time, patients with TBI remained likely to have preventable shortcomings in care (18).
+Methods
+-------
 
-## Aim
+Results
+-------
 
-As such, this study aims to identify and describe opportunities for improvement in the care of adult trauma patients with TBI admitted to the ICU, and to examine patient- and process-related factors associated with their occurrence.
+Conclusion
+----------
+
+Introduction
+============
+
+<!-- Explain the scientific background and rationale for the
+investigation being reported. End by stating the aim or specific
+objectives, including any prespecified hypotheses. The introduction
+should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
+6-8 lines long. Support every statement with a citation -->....
+
+Methods 
+=======
+
+Study design 
+------------
+
+<!-- Present key elements of study design -->
+
+Setting
+-------
+
+<!-- Describe the setting, locations, and relevant dates, including
+periods of recruitment, exposure, follow-up, and data collection -->
+
+Participants
+------------
+
+<!-- Cohort study: Give the eligibility criteria, and the sources and
+methods of selection of participants. Describe methods of
+follow-up. For matched studies, give matching criteria and number of
+exposed and unexposed -->
+
+<!-- Case-control study: Give the eligibility criteria, and the
+sources and + methods of case ascertainment and control
+selection. Give the rationale for the choice of cases and
+controls. For matched studies, give matching criteria and the number
+of controls per case -->
+
+<!-- Cross-sectional study: Give the eligibility criteria, and the
+sources and methods of selection of participants -->
+
+Variables and data sources/measurements
+---------------------------------------
+
+<!-- Clearly define all outcomes, exposures, predictors, potential
+confounders, and effect modifiers. Give diagnostic criteria, if
+applicable. For each variable of interest, give sources of data and
+details of methods of assessment (measurement). Describe comparability
+of assessment methods if there is more than one group -->
+
+Bias 
+----
+
+<!-- Describe any efforts to address potential sources of bias -->
+
+Study size 
+----------
+
+<!-- Explain how the study size was arrived at -->
+
+Quantitative variables
+----------------------
+
+<!-- Explain how quantitative variables were handled in the
+analyses. If applicable, describe which groupings were chosen and why
+-->
+
+Statistical methods
+-------------------
+
+<!--
+
+(a) Describe all statistical methods, including those used to control
+for confounding
+
+(b) Describe any methods used to examine subgroups and interactions
+
+(c) Explain how missing data were addressed
+
+(d)
+
+Cohort study: If applicable, explain how loss to follow-up was addressed
+
+Case-control study:If applicable, explain how matching of cases and
+controls was addressed
+
+Cross-sectional study: If applicable, describe analytical methods
+taking account of sampling strategy
+
+(e) Describe any sensitivity analyses
+-->
+
+Results
+=======
+
+<!--
+
+Participants
+------------
+
+(a) Report numbers of individuals at each stage of study—eg numbers
+potentially eligible, examined for eligibility, confirmed eligible,
+included in the study, completing follow-up, and analysed
+
+(b) Give reasons for non-participation at each stage
+
+(c) Consider use of a flow diagram
+
+Descriptive data
+----------------
+
+(a) Give characteristics of study participants (eg demographic,
+clinical, social) and information on exposures and potential
+confounders
+
+(b) Indicate number of participants with missing data for each
+variable of interest
+
+(c) Cohort study — Summarise follow-up time (eg, average and total
+amount)
+
+Outcome data
+------------
+
+Cohort study — Report numbers of outcome events or summary measures
+over time
+
+Case-control study — Report numbers in each exposure category, or
+summary measures of exposure
+
+Cross-sectional study — Report numbers of outcome events or summary
+measures
+
+Main results
+------------
+
+(a) Give unadjusted estimates and, if applicable, confounder-adjusted
+estimates and their precision (eg, 95% confidence interval). Make
+clear which confounders were adjusted for and why they were included
+
+(b) Report category boundaries when continuous variables were
+categorized
+
+(c) If relevant, consider translating estimates of relative risk into
+absolute risk for a meaningful time period
+
+Other analyses 
+--------------
+
+Report other analyses done—eg analyses of subgroups and interactions,
+and sensitivity analyses 
+-->
+
+You can include code in this document like this:
+
+```{r main, echo=FALSE}
+source("main.R") ## This "imports" the main script file of your project and run any code in it
+```
+
+You can also embed plots:
+
+```{r plot, echo=FALSE}
+plot(pressure)
+```
+
+You can also mix text and code, so called inline code, like this: `r 2+5`.
+
+Discussion
+==========
+
+<!--
+
+Key results
+-----------
+Summarise key results with reference to study objectives
+
+Limitations
+-----------
+
+Discuss limitations of the study, taking into account sources of
+potential bias or imprecision.  Discuss both direction and magnitude
+of any potential bias
+
+Interpretation
+--------------
+
+Give a cautious overall interpretation of results considering
+objectives, limitations, multiplicity of analyses, results from
+similar studies, and other relevant evidence
+
+Generalisability
+----------------
+
+Discuss the generalisability (external validity) of the study results
+
+-->
+
+Conclusion
+==========
+
+<!-- Give a short conclusion, corresponding to the aim -->
+
+References
+==========
+
+<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2025-10-05 functions/OR_Gender.R

@@ -1,0 +1,10 @@
+install.packages("broom")
+
+Data.YN.ofi <- Data.YN.ofi %>% dplyr::mutate(
+    Gender = dplyr::recode(Gender, "K" = "Female", "M" = "Male"), #Changing the variable name for tables and graphs later
+    Gender = factor(Gender, levels = c("Female","Male"))  # Female = reference so the OR will be males vs females
+    )
+
+m_gender <- glm(ofi ~ Gender, data = Data.YN.ofi, family = binomial()) #Performing the simple logistical regression comparing OFI in males Vs females
+
+broom::tidy(m_gender, conf.int = TRUE, exponentiate = TRUE) #Converting log-odds to OR and adding CI

### 2025-10-05 functions/Only_ofi_function.R

@@ -1,10 +1,0 @@
-library(dplyr)
-
-
-filter_ofi <- function(x) 
-{
-  dplyr::filter(x, !is.na(ofi) & ofi != "No") #Filtering patients so that only those where ofi is +ve are left
-  }
-
-ofi.data <- filter_ofi(merged.data) #testing if function works 
-

### 2025-10-05 functions/Remove_Missing_OFI.R

@@ -1,0 +1,7 @@
+library(dplyr)
+
+Remove_missing_ofi <- function(x)  {
+  dplyr::filter(x, !is.na(ofi)) #Filtering patients so that only those with an outcome are left 
+}
+
+Data.YN.ofi <- Remove_missing_ofi(merged.data) #Removing patients that do not have data on OFI 

### 2025-10-05 main.R

@@ -32,5 +32,8 @@ merged.data <- merge_data(data, test = TRUE)
 # Add opportunities for improvement outcome
 merged.data$ofi <- create_ofi(merged.data)
 
-# 
+#Making OFI a factor and not characters 
+merged.data <- merged.data %>%
+  dplyr::mutate(ofi = factor(ofi, levels = c("Yes","No"))) 
+
 

### 2025-10-10 project-plan.qmd
Changing definition of GCS

@@
 ---
-title: "project-plan"
+title: "Project-plan"
 date: today 
@@
 ------------
-
 Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
@@
 ------------
-
-<!-- MGW: Consider discussing the management of trauma first and then the management of TBI, as many of the things you write below also apply to non-TBI trauma patients -->
+<!-- MGW: Consider discussing the management of trauma first and then the management of TBI, as many of the things you
+write below also apply to non-TBI trauma patients -->
 
@@
 ----------------
+<!-- Write that there will be descripitve analysis for OFI prevelance...-->
 

### 2025-10-10 functions/OR_Gender.R

@@ -4,7 +4,6 @@ Data.YN.ofi <- Data.YN.ofi %>% dplyr::mutate(
     Gender = dplyr::recode(Gender, "K" = "Female", "M" = "Male"), #Changing the variable name for tables and graphs later
     Gender = factor(Gender, levels = c("Female","Male"))  # Female = reference so the OR will be males vs females
     )
-
 m_gender <- glm(ofi ~ Gender, data = Data.YN.ofi, family = binomial()) #Performing the simple logistical regression comparing OFI in males Vs females
 
 broom::tidy(m_gender, conf.int = TRUE, exponentiate = TRUE) #Converting log-odds to OR and adding CI

### 2025-10-10 functions/Remove_Missing_OFI.R

@@ -1,4 +1,4 @@
-library(dplyr)
+library(dplyr) #Probs run this function in the main R 
 
 Remove_missing_ofi <- function(x)  {
   dplyr::filter(x, !is.na(ofi)) #Filtering patients so that only those with an outcome are left 

### 2025-10-10 main.R

@@ -36,4 +36,9 @@ merged.data$ofi <- create_ofi(merged.data)
 merged.data <- merged.data %>%
   dplyr::mutate(ofi = factor(ofi, levels = c("Yes","No"))) 
 
+#Filter data to variables that should be observed, remove patients with missing data 
+#Make first descriptive table 
+
+
+
 

### 2025-10-11 functions/TBI_Definer.R
Declaring if the patients have TBI or Not and adding it to the data

@@ -1,0 +1,31 @@
+
+AIS_columns <- grep("^AISCode_", names(merged.data), value = TRUE) #Identifying the columns/variables that have the AIS Code (String)
+
+Is_TBI_AIS <- function(AIS_code) {
+  
+AIS_code <- str_trim(as.character(AIS_code)) #Removing spaces and ensuring that the string is only characters
+
+Ais_empty <- is.na(AIS_code) | AIS_code == "" 
+  
+Body_region <- str_detect(AIS_code, "^1")   # First digit = 1 = Head
+
+Severity_scale <- as.numeric(str_extract(AIS_code, "([0-6])$"))  # Last digit = severity (1–6)
+
+result_ais <- Body_region & !is.na(Severity_scale) & Severity_scale >= 3 #So if the body region is head (1) and the severity is 3 or more then the AIS is according to our def of TBI
+
+result_ais [Ais_empty] <- FALSE #If the AIS code is missing or empty return FALSE and skip going through this string
+
+result_ais
+}
+
+
+
+Is_TBI_GCS <- function (ed_gcs, pre_gcs) { #Function that sees if the patients have had a gcs 8 or under pre hosp or in the ed
+  
+ed_is_TBI_GCS  <- suppressWarnings(as.numeric(ed_gcs))  <= 8 #Converts the value in ed_gcs to a number and hides any warnings (which become NA)
+
+pre_is_TBI_GCS <- suppressWarnings(as.numeric(pre_gcs)) <= 8 # ""
+
+(!is.na(ed_is_TBI_GCS)  & ed_is_TBI_GCS ) | ( !is.na(pre_is_TBI_GCS) & pre_is_TBI_GCS )#If either the ED or pre GCS is 8 or under return true
+  
+}

### 2025-10-11 main.R

@@ -39,6 +39,11 @@ merged.data <- merged.data %>%
 #Filter data to variables that should be observed, remove patients with missing data 
 #Make first descriptive table 
 
+merged.data <- merged.data %>%
+  mutate(
+    TBI = (if_any(all_of(AIS_columns), ~ Is_TBI_AIS(.))) & Is_TBI_GCS(ed_gcs_sum, pre_gcs_sum)
+  )
+
 
 
 

### 2025-10-13 functions/TBI_Definer.R
Comments + starting to code for the function that filters according to my variables

@@ -1,4 +1,5 @@
 
+
 AIS_columns <- grep("^AISCode_", names(merged.data), value = TRUE) #Identifying the columns/variables that have the AIS Code (String)
 
 Is_TBI_AIS <- function(AIS_code) {
@@ -19,7 +20,6 @@ result_ais
 }
 
 
-
 Is_TBI_GCS <- function (ed_gcs, pre_gcs) { #Function that sees if the patients have had a gcs 8 or under pre hosp or in the ed
   
 ed_is_TBI_GCS  <- suppressWarnings(as.numeric(ed_gcs))  <= 8 #Converts the value in ed_gcs to a number and hides any warnings (which become NA)
@@ -29,3 +29,5 @@ pre_is_TBI_GCS <- suppressWarnings(as.numeric(pre_gcs)) <= 8 # ""
 (!is.na(ed_is_TBI_GCS)  & ed_is_TBI_GCS ) | ( !is.na(pre_is_TBI_GCS) & pre_is_TBI_GCS )#If either the ED or pre GCS is 8 or under return true
   
 }
+
+

### 2025-10-13 functions/Variable_filter.R

@@ -1,0 +1,31 @@
+
+#Variables
+
+Variables_wanted <- c(
+
+#Categorical 
+  "Gender", "pt_Gender",                                                         #Sex
+  "",
+#ON call times -> hours outside of 08:00 - 17:00, or Saturday and Sunday
+#Hospital care level
+#Intubation status
+#MEchanism of injury 
+#ASA class
+#Duration of mechanical ventialtion
+#ICU length stay 
+
+#Continous 
+  
+#Age
+#ISS, NISS
+#RTS
+  "pre_sbp_value", "pre_sbp_rtscat", "ed_sbp_value", "ed_sbp_rtscat",           #SBP 
+  "pre_rr_value", "pre_rr_rtscat", "ed_rr_value", "ed_rr_rtscat",               #RR
+  "pre_gcs_sum", "pre_gcs_motor", "ed_gcs_sum", "ed_gcs_motor", 
+  "VK_gcs_less9_ej_intubTE",                                                    #GCS
+  
+#Time to first CT
+#RTS
+
+
+)

### 2025-10-13 main.R

@@ -39,11 +39,17 @@ merged.data <- merged.data %>%
 #Filter data to variables that should be observed, remove patients with missing data 
 #Make first descriptive table 
 
+#Adding column to data that indicates whether the patient has OFI or not 
 merged.data <- merged.data %>%
   mutate(
     TBI = (if_any(all_of(AIS_columns), ~ Is_TBI_AIS(.))) & Is_TBI_GCS(ed_gcs_sum, pre_gcs_sum)
-  )
+  )#How should i handle missing values of either GCS or AIS. They will return as False in the TBI column, maybe does not matter? 
+#I am thinking for my flow chart in results where i filter the patients, do i need to show how many had missing data or maybe enough to show how many had TBI
 
+#Making dataset with only patients that have TBI
+TBI.only.data <- merged.data %>% filter(TBI)
+
+#Cant find DOA as a variable, but filter the data so that they are excluded 
 
 
 

### 2025-11-13 manuscript.qmd
Adding flowchart to results

@@
 
-<!--
-
 Participants
@@
 
-(a) Report numbers of individuals at each stage of study—eg numbers
+<!-- (a) Report numbers of individuals at each stage of study—eg numbers
 potentially eligible, examined for eligibility, confirmed eligible,
@@
 
-(c) Consider use of a flow diagram
+(c) Consider use of a flow diagram -->
+
+ ```{r main, echo=FALSE}
+library(DiagrammeR)
+library(glue)
+source("main.R") ## This "imports" the main script file of your project and run any code in it
+
+# Flowchart figure -----------------------------------------------------------
+
+## Labels for figure ---------------------------------------------------------
+
+label_A <- glue("Trauma registry 2012–2023\n{_n_total_merged_data} patients")
+label_B <- glue("Identified TBI using AIS 1xxxxx.x\n{n_tbi_TBI_only_data} patients with TBI")
+label_C <- glue("Excluded:\n<18 years: {n_under18}\nDead on arrival: {n_doa}")
+label_D <- glue("Linked with trauma care quality database\nExcluded: no OFI outcome: {n_no_ofi}")
+label_E <- glue("Complete-case selection\nExcluded: missing data in covariates: {n_missing_cc}")
+label_F <- glue("Final analytic cohort\n{n_final} adult TBI patients\nOFI: {n_ofi} ({round(100*n_ofi/n_final,1)}%) | No OFI: {n_final - n_ofi}")
+
+## Render flowchart ----------------------------------------------------------
+
+grViz(glue("
+digraph TBI_flow {{
+
+graph [rankdir = TB]
+
+node [shape = box, 
+fontname = 'Times New Roman', 
+fontsize = 12, 
+width = 3]
+
+A [label = '{label_A}']
+B [label = '{label_B}']
+C [label = '{label_C}']
+D [label = '{label_D}']
+E [label = '{label_E}']
+F [label = '{label_F}']
+
+A -> B -> C -> D -> E -> F
+}}
+"))
+
+```
+
+<!--
 

### 2025-11-13 project-plan.qmd
Clarifying TBI def
Adding flowchart to results

@@
 suggests that, despite overall declines in OFI, patients with TBI remain at risk of preventable shortcomings,
-underscoring the need for TBI-specific quality improvement efforts [@Szolnoky2025]. <!--CP: Therefore we hypothesis ze that is possible to identify new OFis related to TBI?? 
-CP: See if something like that can be a hypothesis -->
+underscoring the need for TBI-specific quality improvement efforts [@Szolnoky2025]. Therefore, we
+hypothesise that despite existing trauma quality improvement measures, new OFIs can be identified in
+the care of adult trauma patients with TBI.
 
@@
 =====
-This study aims to identify and describe OFI in the care of adult trauma patients with TBI, and to examine patient- and process-related factors associated with their occurrence.
-<!-- CP: You need to phrase these aims in a more concrete way: this what this course demands..see also next comment E.g.
-(but for your supervisor to help with) Identify specific OFI by doing what?
-I mean , an identification require that certain parameters will be used and confirmed
-If too long M&M will describe these parameters)” -->
-
-<!-- CP: Pls describe this aim in a Yes/No question e..g determine whether the identified OFI are associated with patient and process related factors. Ask for your supervisor to help revising -->
-
-<!--CP: Can you explain more concretely What this means? Occurrence of OPI?-->
+As such, this study aims to determine whether OFIs occur in the care of adult trauma patients with
+TBI, and, if so, to quantify their frequency, describe their characteristics, and examine whether
+patient- and process-related factors are associated with their occurrence.
 
@@
 2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as a Glasgow Coma
-Scale (GCS) score of ≤8 combined with an Abbreviated Injury Scale (AIS) ≥3 for the head region. Patients were
-eligible irrespective of level of care and were not restricted to intensive care admissions.
+Scale (GCS) score of ≤8 combined with an Abbreviated Injury Scale (AIS) severity ≥3 for the head region.
+Patients were eligible irrespective of level of care and were not restricted to intensive care admissions.
 

### 2025-11-13 main.R
Adding comments

@@ -49,6 +49,7 @@ merged.data <- merged.data %>%
 TBI.only.data <- merged.data %>% filter(TBI)
 
 # Cant find DOA as a variable, but filter the data so that they are excluded 
+# Also filter patients <18 years 
 
 TBI.only.data <- TBI.only.data %>%
   Add_OnCall(datetime_col = "DateTime_ArrivalAtHospital") %>%

### 2025-11-17 manuscript.qmd
Producing Flowchart for participants

--- 2025-10-05 manuscript.qmd
@@
 title: "Manuscript title here"
-output: html_document
+format: html
 author: "Your name here"
@@
 
-Abstract
-========
+# Abstract
+
+Hello
 
@@
 
-Background
-----------
+## Background
 
-Methods
--------
+## Methods
 
-Results
--------
+## Results
 
-Conclusion
-----------
+## Conclusion
 
-Introduction
-============
+# Introduction
 
@@
 should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
-6-8 lines long. Support every statement with a citation -->....
+6-8 lines long. Support every statement with a citation -->
 
-Methods 
-=======
+# Methods
 
-Study design 
-------------
+## Study design
 
@@
 
-Setting
--------
+## Setting
 
@@
 
-Participants
-------------
+## Participants
 
@@
 
-Variables and data sources/measurements
----------------------------------------
+## Variables and data sources/measurements
 
@@
 
-Bias 
-----
+## Bias
 
@@
 
-Study size 
-----------
+## Study size
 
@@
 
-Quantitative variables
-----------------------
+## Quantitative variables
 
@@
 
-Statistical methods
--------------------
+## Statistical methods
 
@@
 
-Results
-=======
+# Results
 
-<!--
+## Participants
 
-Participants
-------------
-
-(a) Report numbers of individuals at each stage of study—eg numbers
+<!-- (a) Report numbers of individuals at each stage of study—eg numbers
 potentially eligible, examined for eligibility, confirmed eligible,
@@
 
-(c) Consider use of a flow diagram
+(c) Consider use of a flow diagram -->
+
+```{r flowchart, echo=FALSE}
+
+library(DiagrammeR)
+library(glue)
+library(tidyverse)
+library(gtsummary)
+library(DiagrammeRsvg)
+library(rsvg)         
+source("main.R") ## This "imports" the main script file of your project and run any code in it
+
+# Flowchart figure -----------------------------------------------------------
+
+# ---- Counts for intermediate steps -----------------------------------------
+n_excl_step1 <- n_total_merged_data - n_TBI_only_data
+n_excl_step2 <- n_under18 + n_no_ofi
+
+# ---- Labels ----------------------------------------------------------------
+label_A <- glue(
+  "Trauma registry 2012–2023\n",
+  "n = {n_total_merged_data}"
+)
+
+label_B <- glue(
+  "Identified TBI using AIS 1xxxxx.x\n",
+  "n = {n_TBI_only_data}"
+)
+
+label_C <- glue(
+  "Eligible adult TBI patients\n",
+  "with OFI outcome\n",
+  "n = {n_eligible}"
+)
+
+label_D <- glue(
+  "Included (final analytic cohort)\n",
+  "n = {n_final} adult TBI patients\n",
+  "- OFI: {n_ofi} ({round(100 * n_ofi / n_final, 1)}%)\n",
+  "- No OFI: {n_final - n_ofi}"
+)
+
+label_excl1 <- glue( 
+  "Excluded (n = {n_excl_step1}):\n",
+  "- No TBI according to AIS code"
+  )
+
+label_excl2 <- glue(
+  "Excluded (n = {n_excl_step2}):\n",
+  "- Age < 18 years: {n_under18}\n",
+  "- No OFI outcome: {n_no_ofi}"
+)
+
+label_excl3 <- glue(
+  "Excluded: missing data\n",
+  "in covariates (n = {n_missing_cc})"
+)
+
+# ---- Producing the figure with exclusion boxes ------------------------------
+
+g <- grViz(glue("
+digraph TBI_flow {{
+
+  graph [
+    rankdir = TB,
+    splines = ortho
+  ]
+
+  node [
+    shape = box,
+    style = rounded,
+    fontname = 'Times New Roman',
+    fontsize = 12,
+    width = 3
+  ]
+
+  edge [
+    arrowhead = normal
+  ]
+
+  // Main vertical line
+  A [label = '{label_A}']
+  B [label = '{label_B}']
+  C [label = '{label_C}']
+  D [label = '{label_D}']
+
+  // Side exclusion boxes
+  E1 [label = '{label_excl1}']
+  E2 [label = '{label_excl2}']
+  E3 [label = '{label_excl3}']
+
+  // Invisible junction nodes on the vertical line
+  J1 [shape=point, width=0.001, label=\"\", style=invis]
+  J2 [shape=point, width=0.001, label=\"\", style=invis]
+  J3 [shape=point, width=0.001, label=\"\", style=invis]
+
+  // Main vertical line
+  A -> J1 [arrowhead=none] 
+  J1 -> B       
+  B -> J2 [arrowhead=none] 
+  J2 -> C 
+  C -> J3 [arrowhead=none]
+  J3 -> D
+
+  // Branching arrows from the vertical line to exclusions
+  J1 -> E1
+  J2 -> E2
+  J3 -> E3
+
+  // Keep exclusions horizontally aligned with the junctions
+  {{rank=same; J1; E1}}
+  {{rank=same; J2; E2}}
+  {{rank=same; J3; E3}}
+
+}}
+"))
+
+# ---- Export (SVG + PNG) ----
+svg_txt <- DiagrammeRsvg::export_svg(g)
+
+dir.create("assets", showWarnings = FALSE)
+
+out_svg <- file.path("assets", "flowchart.svg")
+out_png <- file.path("assets", "flowchart.png")
+
+cat(svg_txt, file = out_svg)
+rsvg_png(out_svg, file = out_png, width = 1600, height = 2600)
+
+knitr::include_graphics(out_png)
+
+```
+
+<!--
 
@@
 
-Discussion
-==========
+# Discussion
 
@@
 
-Conclusion
-==========
+# Conclusion
 
@@
 
-References
-==========
+# References
 

### 2025-11-17 main.R

@@ -6,16 +6,19 @@ library(rofi) # installed using remotes::install_github("martingerdin/rofi")
 # --- Organising raw data ------------------------------------------
 
 # Import data
-data <- import_data()
+data <- import_data(test = TRUE)
 
 # Merge data and remove duplicate variables/columns
-merged.data <- merge_data(data)
+merged.data <- merge_data(data, test = TRUE)
 merged.data <- merged.data[, !duplicated(names(merged.data))]
 
 # Add opportunities for improvement outcome and making it a factor
 merged.data$ofi <- create_ofi(merged.data)
 merged.data <- merged.data %>%
   dplyr::mutate(ofi = factor(ofi, levels = c("No","Yes"))) 
+
+n_total_merged_data <- nrow(merged.data)
+
 
 # --- Adding TBI to data --------------------------------------------------
 
@@ -47,6 +50,7 @@ merged.data <- merged.data %>%
 
 # Making dataset with only patients that have TBI
 TBI.only.data <- merged.data %>% filter(TBI)
+n_TBI_only_data <- sum(TBI.only.data$TBI)
 
 # Cant find DOA as a variable, but filter the data so that they are excluded 
 # Also filter patients <18 years 
@@ -85,8 +89,16 @@ Variables_wanted <- c(
 # Filtering my dataset to only the variables i want, including RTS and On call. How they are derived can be seen in Variable_filter
 TBI.only.filtered <- TBI.only.data %>% select(all_of(Variables_wanted))
 
-# Removing patients with missing Ofi, not really final sample because for my regression i will be removing all patients with missing 
-Analysis.sample <- Remove_missing_ofi(TBI.only.filtered) 
+# Excluding individuals that are under 18
+n_under18 <- sum(TBI.only.filtered$pt_age_yrs < 18) 
+
+TBI.only.filtered <- TBI.only.filtered %>% 
+  filter(pt_age_yrs >= 18)
+
+# Removing patients with missing Ofi
+n_no_ofi <- sum(is.na(TBI.only.filtered$ofi))
+
+Analysis.sample <- Remove_missing_ofi(TBI.only.filtered)
 
 # --- Descriptive table -------------------------------------------
 
@@ -163,7 +175,7 @@ Descriptive.table1 # First baseline descriptive table
 
[…]

### 2025-11-17 manuscript.qmd
Flowchart

@@
 
-dir.create("assets", showWarnings = FALSE)
-
 out_svg <- file.path("assets", "flowchart.svg")
@@
 
-knitr::include_graphics(out_png)
-
 ```
 
+![**Figure 1. Flowchart of patient selection for the analytical cohort**](assets/flowchart.png)
+
+## Descriptive data
+
 <!--  
-
-Descriptive data
-----------------
-
 (a) Give characteristics of study participants (eg demographic,
@@
 amount)
+-->
+## Outcome data
 
-Outcome data
-------------
-
+<!--  
 Cohort study — Report numbers of outcome events or summary measures
@@
 measures
+-->
 
-Main results
-------------
+##  Main results
 
+<!--  
 (a) Give unadjusted estimates and, if applicable, confounder-adjusted
@@
 absolute risk for a meaningful time period
+-->
 
-Other analyses 
---------------
+## Other analyses
 
+<!--  
 Report other analyses done—eg analyses of subgroups and interactions,
@@
 
+## Key results
+
 <!--
+Summarise key results with reference to study objectives
+-->
 
-Key results
------------
-Summarise key results with reference to study objectives
+## Limitations
 
-Limitations
------------
-
+<!--
 Discuss limitations of the study, taking into account sources of
@@
 of any potential bias
+-->
 
-Interpretation
---------------
+## Interpretation
 
+<!--
 Give a cautious overall interpretation of results considering
@@
 similar studies, and other relevant evidence
+-->
 
-Generalisability
-----------------
+## Generalisability
 
+<!--
 Discuss the generalisability (external validity) of the study results
-
 -->

### 2025-11-18 main.R
Adding tables to manuscript

@@ -6,10 +6,10 @@ library(rofi) # installed using remotes::install_github("martingerdin/rofi")
 # --- Organising raw data ------------------------------------------
 
 # Import data
-data <- import_data(test = TRUE)
+data <- import_data()
 
 # Merge data and remove duplicate variables/columns
-merged.data <- merge_data(data, test = TRUE)
+merged.data <- merge_data(data)
 merged.data <- merged.data[, !duplicated(names(merged.data))]
 
 # Add opportunities for improvement outcome and making it a factor

### 2025-11-19 functions/TBI_Definer.R
Correct comment

@@ -14,6 +14,6 @@ Get_TBI_Severity <- function(AIS_code) {
   AIS.sev <- suppressWarnings(as.numeric(str_extract(AIS_code, "([0-6])$")))
   if (all(is.na(AIS.sev))) return(NA_real_)
   
-  # Return the most severe/max AIS value among the patient's 140-codes.
+  # Return the most severe/max AIS value among the patient's 1-codes.
   max(AIS.sev, na.rm = TRUE)
 }

### 2025-12-10 manuscript.qmd
Adding tables to manuscript
Testing version control
Writing abstract + intro for manuscript

@@
 ---
-title: "Manuscript title here"
+title: "Opportunities for improvement in the care of adult patients with traumatic brain injury"
 format: html
-author: "Your name here"
+author: "Vivek Kanwar"
 bibliography: bibliography.bib
@@
 # Abstract
-
-Hello
-
 <!-- Provide in the abstract an informative and balanced summary of
 what was done and what was found. Not more than 300 words. -->
+## Background
 
-## Background
+Traumatic brain injury (TBI) is a major cause of death and disability after trauma.
+Trauma quality programmes increasingly use “opportunities for improvement” (OFI) to identify deviations from optimal care, but patterns of OFI in adult TBI are not well described.
+As such, this study aims to determine if and how often OFIs occur in the care of adult patients with TBI, describe their main characteristics, and examine whether patient- and process-related factors are associated with OFI
 
@@
 
+A registry-based retrospective cohort study at Karolinska University Hospital in Solna, Sweden, centre using the local trauma registry (2012–2023) linked to the hospital’s trauma care quality database was conducted.
+Adult TBI patients were identified using the Abbreviated Injury Scale (AIS).
+The outcome was the presence of at least one OFI, classified as yes/no through a standardised multidisciplinary review.
+Patient- and process-related variables were analysed in complete-case logistic regression models with OFI status as the dependent variable.
+
 ## Results
 
+OFIs were identified in 7.7% of TBI patients.
+In adjusted analyses, higher ISS (adjusted OR 1.05, 95% CI 1.03–1.07), higher Revised Trauma Score (adjusted OR 1.32, 95% CI 1.03–1.72; p = 0.034), admission to a high-dependency unit compared to the emergency ward (adjusted OR 3.50, 95% CI 1.30–11.13) and longer hospital stay (adjusted OR 1.02 per day, 95% CI 1.00–1.04) were independently associated with OFI.
+
 ## Conclusion
+
+In this adult TBI cohort, OFIs were relatively uncommon but found in patients with high anatomical injury burden, seemingly stable physiology, intermediate care levels and longer hospital stays.
+These patterns suggest that trauma quality improvement efforts may be particularly relevant in TBI patients with complex care trajectories and those who appear physiologically stable despite substantial injury burden.
 
@@
 6-8 lines long. Support every statement with a citation -->
+Traumatic injury is a major global health problem and a leading cause of death and disability, particularly among people younger than 45 years [@Maerz2009].
+Within this burden, traumatic brain injury (TBI) accounts for a large proportion of trauma-related mortality and long-term neurological impairment and has been highlighted as a priority area for trauma systems (2,3).
+TBI arises when external forces causes acute neurological impairment and initiate both primary structural damage and secondary pathophysiological cascades, including oedema, ischemia and metabolic disturbances which develop over time and critically influence long-term outcomes (4,5).
+
+## Management of TBI
+
+After initial stabilisation, management of TBI focuses on preventing secondary brain injury by securing the airway, ensuring adequate ventilation and rapidly correcting hypotension, as lapses in these areas are associated with increased mortality (6-9).
+In terms of imaging, rapid non-contrast head CT is essential to detect surgically liable lesions, while in patients with multi-system trauma, whole-body CT may be considered (10,11).
+International guidelines and reviews, including those from the Brain Trauma Foundation (BTF), provide recommendations on monitoring, control of intracranial- and cerebral perfusion pressure, and avoidance of potentially harmful interventions such as corticosteroids (5,7,13-15).
+Despite these frameworks, the heterogeneous and dynamic nature of TBI makes care particularly susceptible to delays, omissions and deviations from optimal practice (5,7,13).
+
+## Trauma Quality Improvement
+
+To address such vulnerabilities in trauma cohorts, trauma quality improvement programmes (TQIPs) systematically evaluate care processes and outcomes.
+These programmes use audit filters, structured case reviews and multidisciplinary morbidity and mortality conferences to find “opportunities for improvement” (OFI), defined as deviations from optimal care that may have contributed to adverse outcomes or  preventable harm (17-19).
+The World Health Organization and the American College of Surgeons provide guidance on how these programmes should be organised, emphasising continuous, multidisciplinary review of trauma cases and feedback to clinicians (18,19).
+
+Previous studies in general trauma populations, and in mature trauma settings even, have shown that deviations and potentially preventable outcomes often arise from shortcomings in areas regarding, airway management, haemorrhage control, resuscitation and delays in imaging or missed injuries (3,20-23).
+Within this framework, patterns focused specifically on TBI is more limited.
+Studies of preventable or potentially preventable TBI deaths report recurrent issues such as delayed neurosurgical intervention, suboptimal early resuscitation and inadequate monitoring, suggesting that TBI may carry distinct vulnerabilities along the trauma pathway (24,25).
+
+## Rationale for this study
+
+TBI remains a challenge in modern system because even small deviations from optimal practice can markedly worsen outcomes.
+Although TQIPs aim to identify and correct such lapses, most research has focused on mixed trauma cohorts, leaving TBI-specific vulnerabilities under-characterised where also recent work suggests that patients with TBI remain at risk of OFIs (26).
+As such, this study aims to determine whether OFIs occur in the care of adult trauma patients with TBI, and, if so, to quantify their frequency, describe their characteristics, and examine whether patient- and process-related factors are associated with their occurrence.
 
@@
 
-library(DiagrammeR)
 library(glue)
-library(tidyverse)
-library(gtsummary)
 library(DiagrammeRsvg)
-library(rsvg)         
 source("main.R") ## This "imports" the main script file of your project and run any code in it
@@
 -->
+```{r}
+Descriptive.table1
+```
 ## Outcome data
@@
 -->
+
+```{r}
+Merged.Table
+```
 
@@
 
-<!--
-Discuss the generalisability (external validity) of the study results
--->
+<!-- Discuss the generalisability (external validity) of the study results -->
 

### 2025-12-10 main.R

@@ -1,6 +1,6 @@
 noacsr::source_all_functions()
 
-# Load packages
+# --- Loading packages ---------------------------------------------------------
 library(rofi) # installed using remotes::install_github("martingerdin/rofi")
 library(gtsummary)
 library(broom)
@@ -332,7 +332,7 @@ and opportunities for improvement in TBI patients**"
   )
 # MV.Table1 # Table describing adjusted associations
 
-# -----Combined unadjusted & adjusted table (Table2) --------------------------
+# ---- Combined unadjusted & adjusted table (Table2) --------------------------
 
 Merged.Table <- tbl_merge(
   tbls = list(SR.Table1, MV.Table1),
